---
title: "Whitney Henry - scRNAseq Analysis - PDAC mouse (Pitter, PMID 35952360) - BH4 Pathway Scoring Figures"
author: "Evelyn Tong"
output:
  html_document:
    toc: true
    toc_depth: 3
  html_notebook: default
date: "9/22/2026"
---

Split out of the original monolithic `tmp/WH_BH4pathway_final/src/Analysis_script.Rmd`
("BH4 analysis script - Evelyn - 2026/09/16", Data1 section) during the
`datasets/` reorg, into its own self-contained dataset directory/project.
See `PDAC_mouse_BH4_scoring_Methods.md` for the full methodology, including
a resolved Known Gaps note on the Extended 6a panel's `p1` (an accidental
deletion in the source script, restored per the author's confirmation).

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, message = FALSE, warning = FALSE,
  fig.align = "center", fig.width = 6, fig.height = 4, dpi = 300
)
source("emt_code/bh4_pathway_shared_setup.R")
```

## parameter settings
```{r}
# figures/ + stats output for this dataset
current.outputfolder <- "figures"
# define where the input RDS file is
data.path <- "pitter_bh4_seurat.rds"
# define state categories and colors
STATE_ORDER <- c("Luminal cells", "Basal cells", "Mesenchymal cells", "Other")
state_color <- c("Luminal cells"     = "#334589",
                 "Basal cells"       = "#ed9d2d",
                 "Mesenchymal cells" = "#d95f02",
                 "Other"             = "#C0C0C0")
```
## load data and processing
```{r}
seu <- readRDS(data.path) #
seu$cell_state <- factor(as.character(seu@meta.data$cell_state), levels = STATE_ORDER)
cat("  Dimensions:", dim(seu)[1], "genes x", dim(seu)[2], "cells\n")
cat("  State counts:\n"); print(table(seu$cell_state, useNA = "ifany"))
cat("  BH4 genes present:", sum(bh4_genes_mouse %in% rownames(seu)), "/", length(bh4_genes_mouse), "\n")

seu <- NormalizeData(seu)

# ── Dataset-customized UCell maxRank ─────────────────────────────────────────
# Malignant compartment = cells used for plotting/correlation (non-"Other").
if (!"nFeature_RNA" %in% colnames(seu@meta.data)) {
  seu$nFeature_RNA <- Matrix::colSums(GetAssayData(seu, assay = "RNA", layer = "counts") > 0)
}
malignant_mask <- !is.na(seu$cell_state) & seu$cell_state != "Other"
ucell_maxrank_custom <- as.integer(round(median(seu$nFeature_RNA[malignant_mask])))

cat(sprintf(
  "\n  >>> Custom UCell maxRank = median nFeature_RNA of malignant compartment\n"))
cat(sprintf(
  "      malignant cells (cell_state != 'Other'): n = %d\n", sum(malignant_mask)))
cat(sprintf(
  "      median nFeature_RNA (chosen maxRank)   : %d\n", ucell_maxrank_custom))
```
## calculate BH4 signature score
```{r}
# Run UCell, maxRank = dataset-customized maxRank (median nFeature of malignant)
seu <- AddModuleScore_UCell(
  seu,
  features = bh4_sigs_mouse,
  maxRank = ucell_maxrank_custom,
  name = "_UCellMedianMaxRank",
  ncores = 4
)

# Composite (full-pathway) scores for the summary / UMAP / heatmap
seu$BH4_UCellMedianMaxRank <- seu@meta.data[[paste0(bh4_full, "_UCellMedianMaxRank")]]
```

## plotting
### Main 5d - UMAP by BH4 score

```{r}
# Extract UMAP coordinates and just the bh4_full score
umap_coords <- as.data.frame(Embeddings(seu, "umap"))[, 1:2]
colnames(umap_coords) <- c("UMAP_1", "UMAP_2")
umap_coords$cell_state <- seu$cell_state
df <- umap_coords[, c("UMAP_1", "UMAP_2")]
df$score <- seu@meta.data[[paste0(bh4_full, "_UCellMedianMaxRank")]]

# Create the single UMAP plot
p_umap_ucellMMR <- ggplot(df, aes(UMAP_1, UMAP_2, color = score)) +
  geom_point(size = 0.2, stroke = 0, shape = 16) +
  scale_color_gradient(low = "grey90", high = "#CB181D", name = "Score", limits = c(0, 0.8), n.breaks = 3) +
  labs(
    title = sprintf("BH4 UCell score (maxRank = %d)", ucell_maxrank_custom),
    x = "UMAP 1",
    y = "UMAP 2"
  ) +
  figure_theme +
  theme(
    axis.text = element_blank(),
    axis.ticks = element_blank(),
    aspect.ratio = 1
  ) +
  coord_fixed()

# Save the figure with 2x2 dimensions
save_fig(p_umap_ucellMMR, "5d_UMAP_BH4_score", 2, 2)
```
### Extended 6a UMAP by BH4 gene expression

`p1` (a narrow UMAP colored by `cell_state`, paired with `p3`'s wider
per-gene expression panel below) was missing from the source script --
the author confirmed this was accidentally deleted and supplied the
original definition, restored verbatim here. See Methods.md.

```{r}
### UMAP by cell state
df_umap <- umap_coords[order(umap_coords$cell_state == "Other", decreasing = TRUE), ]
p1 <- ggplot(df_umap, aes(UMAP_1, UMAP_2, color = cell_state)) +
  geom_point(size = 0.2, stroke = 0, shape = 16) +
  scale_color_manual(values = state_color, name = "Cell state", drop = FALSE) +
  labs(title = "Autochthonous KPCT cells", subtitle = paste0(nrow(df_umap), " cells"),
       x = "UMAP 1", y = "UMAP 2") +
  figure_theme +
  theme(axis.text = element_blank(), axis.ticks = element_blank(), aspect.ratio = 1) +
  guides(color = guide_legend(override.aes = list(size = 1.2))) +
  coord_fixed()

### BH4 genes
data_mat       <- GetAssayData(seu, assay = "RNA", layer = "data")
bh4_lognorm    <- as.matrix(t(data_mat[bh4_genes_mouse, ]))
colnames(bh4_lognorm) <- bh4_genes_mouse

for (g in bh4_genes_mouse) df[[g]] <- bh4_lognorm[, g]
umap_long <- df %>%
  pivot_longer(cols = all_of(bh4_genes_mouse), names_to = "gene", values_to = "expression") %>%
  mutate(gene = factor(gene, levels = bh4_genes_mouse, labels = bh4_genes_mouse))
p3 <- ggplot(umap_long, aes(UMAP_1, UMAP_2, color = expression)) +
  geom_point(size = 0.2, stroke = 0, shape = 16) +
  facet_wrap(~ gene, nrow = 1) +
  scale_color_gradient(low = "grey90", high = "#CB181D", name = "Expr.") +
  labs(title = "BH4 gene expression (log-normalized)", x = "UMAP 1", y = "UMAP 2") +
  figure_theme +
  theme(axis.text = element_blank(), axis.ticks = element_blank(),
        strip.text = element_text(size = 5, face = "bold"), strip.background = element_blank(),
        aspect.ratio = 1) +
  coord_fixed()

p_combined <- plot_grid(p1, p3, rel_widths = c(1.5, 6))
save_fig(p_combined, "Extended_6a_UMAP_byCellState_GeneExpression", 10, 1.5)
```
### Main 5e boxplot by BH4 score

```{r}
box_long_scores <- do.call(rbind, lapply(bh4_score_names, function(s) {
    data.frame(
        Quantity = s,
        cancer_state = seu$cell_state,
        value = as.numeric(seu@meta.data[[paste0(s, "_UCellMedianMaxRank")]]),
        y_label = "Signature score",
        stringsAsFactors = FALSE
    )
})) %>% subset(cancer_state != "Other")

title_str_scores <- sprintf("BH4 signatures by cell state  |  UCell (maxRank = %d)", ucell_maxrank_custom)
out_file_scores  <- file.path(current.outputfolder, "5e_Boxplot_byCellState_BH4_score.pdf")

make_boxplot_state(
    box_long_scores,
    bh4_score_names,
    title_str_scores,
    out_file_scores,
    levels = STATE_ORDER[1:3],
    fills = state_color[1:3]
)
```

### Extended 6b6c boxplot by BH4 gene expression

```{r}
box_long_genes <- do.call(rbind, lapply(bh4_genes_mouse, function(g) {
    data.frame(
        Quantity = g,
        cancer_state = seu$cell_state,
        value = as.numeric(data_mat[g, ]),
        y_label = "Log-normalized expression",
        stringsAsFactors = FALSE
    )
})) %>% subset(cancer_state != "Other")

title_str_genes <- "BH4 genes by cell state"
out_file_genes  <- file.path(current.outputfolder, "Extended_6c_Boxplot_byCellState_GeneExpression.pdf")

make_boxplot_state(
    box_long_genes,
    bh4_genes_mouse,
    title_str_genes,
    out_file_genes,
    levels = STATE_ORDER[1:3],
    fills = state_color[1:3]
)
```

## write session info

```{r session-info}
sessionInfo()
writeLines(capture.output(sessionInfo()), "PDAC_mouse_BH4_scoring_sessionInfo.txt")
```
