---
title: "Whitney Henry - SUM149PT analysis - BH4 and other set scores"
author: "Charlie Whittaker"
output:
  html_document:
    toc: true
    toc_depth: 3
  html_notebook: default
date: "8/31/2026"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
set.seed(1234)
```

```{r load-packages, message=FALSE, warning=FALSE}
library(openxlsx)
library(tidyverse)
library(reprex)
library(matrixStats)
library(XML)
library(ggrepel)
library(rtracklayer)
library(DESeq2)
library(apeglm)
library(multcomp)
library(ComplexHeatmap)
library(tximport)
library(edgeR)
library(gprofiler2)
library(cluster)
library(stringr)
library(fgsea)
library(GSVA)
library(ggpubr)
library(Seurat)
library(UCell)
```

# Goal

Score `protCoding`'s 18 SUM149PT EMT-spectrum samples (GSE172609: E, EM1,
EM2, EM3, M1, M2; 3 reps each, per `Sample_to_Condition.xlsx` -- the 3
`Parental` reps are excluded when `protCoding` is built) for
EMT status (KS) and pathway/gene-set activity (GSVA, ssGSEA, UCell -- BH4 +
subsets, Glutathione, CoQ10, Thioredoxin, Peroxiredoxin, and the Epi/Mes KS
signature genes themselves), then plot those scores across the experimental
EMT spectrum. Scoring methodology (KS - for EMT spectrum, GSVA/ssGSEA/Ucell for gene set scoring).
Grouping uses the samples' own annotated `EMT_State` because this dataset already has 
ground-truth EMT-spectrum labels.

# Load rnaseq data from geo

Data from Brown, M.S., Abdollahi, B., Wilkins, O.M. et al. Phenotypic
heterogeneity driven by plasticity of the intermediate EMT state governs
disease progression and metastasis in breast cancer. *Sci. Adv.* 8, eabj8002
(2022). <https://doi.org/10.1126/sciadv.abj8002>. Downloaded from GEO
accession GSE172609 (RNA-seq SubSeries of the study's SuperSeries,
GSE172613).

```{r load-geo-data}
expressionData <- read.delim("from_geo/GSE172609_norm_counts_TPM_GRCh38.p13_NCBI.tsv")
countData <- read.delim("from_geo/GSE172609_raw_counts_GRCh38.p13_NCBI.tsv")
annotationData <- read.delim("from_geo/Human.GRCh38.p13.annot.tsv")
annotationData <- annotationData %>% dplyr::select(GeneID,Symbol,GeneType)
emt.state <- read.xlsx("Sample_to_Condition.xlsx")
```

## Prepare gene expression matrix for scoring

### Annotate the data

```{r}
# --- Store pre-join dimensions for reporting ---
nrow_before <- nrow(expressionData)
n_annot     <- nrow(annotationData)

# --- Join annotationData to expressionData ---
# left_join keeps all rows in expressionData; Symbol/GeneType columns are added
expressionData.annotated <- expressionData %>%
  left_join(annotationData, by = "GeneID")

# --- Move Symbol and GeneType columns to front for readability ---
expressionData.annotated <- expressionData.annotated %>%
  relocate(GeneID, Symbol, GeneType)

# ============================================================
# Join Success Report
# ============================================================
nrow_after       <- nrow(expressionData.annotated)
n_matched        <- sum(!is.na(expressionData.annotated$Symbol))
n_unmatched      <- sum(is.na(expressionData.annotated$Symbol))
n_annot_unmatched <- sum(!annotationData$GeneID %in% expressionData$GeneID)

cat("=== Join Report ===\n")
cat(sprintf("  Rows in expressionData (before join) : %d\n", nrow_before))
cat(sprintf("  Rows in annotationData               : %d\n", n_annot))
cat(sprintf("  Rows in joined dataframe             : %d\n", nrow_after))
cat(sprintf("  Row count preserved                  : %s\n", ifelse(nrow_before == nrow_after, "YES", "NO")))
cat(sprintf("  GeneIDs successfully annotated       : %d\n", n_matched))
cat(sprintf("  GeneIDs with NO annotation (NA)      : %d\n", n_unmatched))
cat(sprintf("  annotationData GeneIDs not in expr.  : %d\n", n_annot_unmatched))
cat("===================\n")

# --- Optional: inspect any unmatched GeneIDs ---
if (n_unmatched > 0) {
  cat("\nGeneIDs with missing annotation:\n")
  print(expressionData.annotated %>% filter(is.na(Symbol)) %>% select(GeneID))
}
```

### Calculate row averages

```{r}
expressionData.annotated <- expressionData.annotated %>% dplyr::mutate(Avg = rowMeans(across(starts_with("GSM")), na.rm = TRUE))
```

### Adding MPG column

Often multiple geneids will map to the same gene symbol. In these cases, it is helpful to select a representative geneID for each group of gene symbols. The code block below iterates through each unique gene symbol, compares all of the geneids associated with the same gene symbol and identifies the geneid with the highest average expression. This gene is then set as the representative (max-per-gene) geneid for that symbol. Note: If multiple geneids have the same highest avg expression, the first gene that appears in the dataframe is taken

The code below performs the following steps:
  1) Create a vector called "mpg" of "No" values the same length as the number of rows in the dataframe assembleData
  2) Loop over the assembleDat object ... identifying index values of duplicate genes, updating mpg vector
  3) Merge mpg vector to assemblDat object with setting column name as "MPG"


```{r}
mpg <- rep('No', nrow(expressionData.annotated))

for (i in unique(expressionData.annotated[,"Symbol"])){
    if (is.na(i)){next}
    idxs <- which(expressionData.annotated[,"Symbol"] == i)
    max_idx <- which(expressionData.annotated$Avg[idxs] == max(expressionData.annotated$Avg[idxs]))[1]

    mpg[idxs[max_idx]] <- 'Yes'
}

expressionData.annotated$MPG <- mpg
```

### Select protein coding genes, one value per gene

The 3 `Parental` ("P") group samples are excluded here -- they are not
carried into any downstream scoring or plots.

```{r}
parental_samples <- emt.state$GEO_Accession[emt.state$EMT_State == "Parental"]

protCoding <- expressionData.annotated %>%
  dplyr::filter(GeneType == "protein-coding" & MPG == "Yes") %>%
  dplyr::select(Symbol,starts_with("GSM")) %>%
  dplyr::select(-all_of(parental_samples))
```

# Gene sets

## Additional gene sets for GSVA/ssGSEA scoring

```{r}
# BH4 synthesis and recycling
BH4 <- c("DHFR", "GCH1", "PCBD1", "PCBD2", "PTS", "QDPR", "SPR")

# De novo BH4 synthesis only
BH4.synth <- c("GCH1", "PTS", "SPR")

# BH4 recycling and regeneration only
BH4.recyc <- c("QDPR", "DHFR", "PCBD1", "PCBD2")

# Cystine uptake and GSH synthesis/recycling
Glutathione <- c("SLC7A11", "SLC3A2", "GCLC", "GCLM", "GSS", "GSR")

# CoQ/Ubiquinol biosynthesis
CoQ10 <- c("COQ2", "COQ3", "COQ4", "COQ5", "COQ6", "COQ7",
            "COQ8A", "COQ8B", "COQ9", "HPDL", "PDSS1", "PDSS2", "STARD7")

# Thioredoxin system
Thioredoxin <- c("TXN", "TXNRD1", "TXN2", "TXNRD2")

# Canonical peroxiredoxins
Peroxiredoxin <- c("PRDX1", "PRDX2", "PRDX3", "PRDX4", "PRDX5")

# Individual genes
goi <- c("GPX4", "AIFM2", "DHODH", "PRDX6")

# Canonical EMT marker genes (used by the MLR method in other notebooks) --
# kept here for reference; not used in any scoring below (KS/GSVA/ssGSEA/UCell)
mlr <- c("CDH1", "VIM", "CLDN7")
```

## KS signature (Epi/Mes) genes

Although SUM149PT is a cultured cell line, not a tumor, the tumor-calibrated
KS signature (`EM_gene_signature_tumor_KS.xlsx`) is used here to match
`TCGA_BRCA_EMT_GeneSet_Scoring.Rmd`'s signature for its tumor samples
(the cell-line-calibrated `EM_gene_signature_cellLine_KS.xlsx` is available
in `emt_code/` if the cell-line signature is wanted instead).

```{r ks-signature-genes}
KSScore.EMTSignature.file <- "../emt_code/EM_gene_signature_tumor_KS.xlsx"

emt.tumor <- read.xlsx("../emt_code/EM_gene_signature_tumor_KS.xlsx", colNames = FALSE)
emt.tumor <- emt.tumor %>% dplyr::rename(gene = X1, group = X2)
Epi.tum <- emt.tumor$gene[emt.tumor$group == "Epi"]
Mes.tum <- emt.tumor$gene[emt.tumor$group == "Mes"]
```

## Combine gene sets for scoring

```{r}
pathways_to_score <- list(
  BH4           = BH4,
  BH4.synth     = BH4.synth,
  BH4.recyc     = BH4.recyc,
  Glutathione   = Glutathione,
  CoQ10         = CoQ10,
  Thioredoxin   = Thioredoxin,
  Peroxiredoxin = Peroxiredoxin,
  Epi.tum        = Epi.tum,
  Mes.tum        = Mes.tum)
```

# Differential expression (DESeq2 + apeglm)

DESeq2 differential expression, `E` set as the reference `EMT_State` level,
each other state (`EM1`, `EM2`, `EM3`, `M1`, `M2`) compared to it. Runs on
raw counts (`countData`), not the `log2(TPM + 1)` matrix used for scoring
above -- DESeq2 does its own size-factor normalization and needs integer
counts. Gene panel mirrors `protCoding` (protein-coding, one row per gene
via the `MPG` selection already computed above) so the same genes are used
for scoring and DE; `Parental` ("P") samples are excluded here too, per
`parental_samples` above. Shrinkage uses `apeglm` (`lfcShrink(..., type =
"apeglm")`), which needs a `coef` (not a `contrast`) -- fine here since
every comparison is already "some level vs the reference level" by
construction of the design/releveling below. The eventual goal is a
heatmap of (shrunk) log2 fold-change values for the BH4 gene set across
the 5 comparisons; this section builds up to a `BH4 x EMT_State` log2FC
matrix and a first-pass heatmap of it.

## Build raw count matrix

```{r deseq-count-matrix}
countData.annotated <- countData %>%
  dplyr::left_join(
    expressionData.annotated %>% dplyr::select(GeneID, Symbol, GeneType, MPG),
    by = "GeneID"
  )

countCoding <- countData.annotated %>%
  dplyr::filter(GeneType == "protein-coding" & MPG == "Yes") %>%
  dplyr::select(Symbol, starts_with("GSM")) %>%
  dplyr::select(-all_of(parental_samples))

deseqCountMat <- countCoding %>%
  tibble::column_to_rownames("Symbol") %>%
  as.matrix() %>%
  round()
storage.mode(deseqCountMat) <- "integer"

dim(deseqCountMat)
```

## Sample metadata

```{r deseq-coldata}
deseqColData <- emt.state %>%
  dplyr::rename(GSM = GEO_Accession) %>%
  dplyr::filter(GSM %in% colnames(deseqCountMat)) %>%
  dplyr::mutate(EMT_State = factor(EMT_State, levels = c("E", "EM1", "EM2", "EM3", "M1", "M2"))) %>%
  tibble::column_to_rownames("GSM")

deseqColData <- deseqColData[colnames(deseqCountMat), , drop = FALSE]

stopifnot(all(rownames(deseqColData) == colnames(deseqCountMat)))
```

## Run DESeq2

```{r deseq-run}
dds <- DESeqDataSetFromMatrix(
  countData = deseqCountMat,
  colData   = deseqColData,
  design    = ~ EMT_State
)
dds$EMT_State <- relevel(dds$EMT_State, ref = "E")

dds <- DESeq(dds)

resultsNames(dds)
```

## Shrink log fold changes (apeglm), each state vs `E`

```{r deseq-apeglm-shrinkage}
emt_state_levels <- setdiff(levels(dds$EMT_State), "E")

deseq_results_list <- purrr::map(emt_state_levels, function(lvl) {
  coef_name <- paste0("EMT_State_", lvl, "_vs_E")
  lfcShrink(dds, coef = coef_name, type = "apeglm")
})
names(deseq_results_list) <- emt_state_levels
```

```{r deseq-results-long}
deseq_results_long <- purrr::map_dfr(names(deseq_results_list), function(lvl) {
  as.data.frame(deseq_results_list[[lvl]]) %>%
    tibble::rownames_to_column("Gene") %>%
    dplyr::mutate(EMT_State = lvl) %>%
    dplyr::select(Gene, EMT_State, log2FoldChange, lfcSE, padj)
}) %>%
  dplyr::mutate(EMT_State = factor(EMT_State, levels = emt_state_levels))
```

## BH4 gene log2FC heatmap

```{r deseq-bh4-lfc-matrix}
bh4_lfc_mat <- deseq_results_long %>%
  dplyr::filter(Gene %in% BH4) %>%
  dplyr::select(Gene, EMT_State, log2FoldChange) %>%
  tidyr::pivot_wider(names_from = EMT_State, values_from = log2FoldChange) %>%
  tibble::column_to_rownames("Gene") %>%
  as.matrix()

bh4_lfc_mat <- bh4_lfc_mat[BH4, emt_state_levels]

bh4_lfc_mat
```

## BH4 gene log2FC heatmap, significance-flagged

An `*` is drawn in any cell that is both a "large" fold-change (|log2FC| >= 1) and statistically
significant (BH-adjusted p <= 0.05).

```{r bh4-lfc-heatmap-sig, fig.width=7, fig.height=6}
bh4_padj_mat <- deseq_results_long %>%
  dplyr::filter(Gene %in% BH4) %>%
  dplyr::select(Gene, EMT_State, padj) %>%
  tidyr::pivot_wider(names_from = EMT_State, values_from = padj) %>%
  tibble::column_to_rownames("Gene") %>%
  as.matrix()

bh4_padj_mat <- bh4_padj_mat[BH4, emt_state_levels]

lfc_col_fun <- circlize::colorRamp2(
  c(min(bh4_lfc_mat, na.rm = TRUE), 0, max(bh4_lfc_mat, na.rm = TRUE)),
  c("#2166AC", "white", "#B2182B")
)

bh4_lfc_heatmap_sig <- Heatmap(
  bh4_lfc_mat,
  name              = "log2FC\n(vs E)",
  col               = lfc_col_fun,
  cluster_columns   = FALSE,
  cluster_rows      = FALSE,
  column_title      = "BH4 gene log2FC vs EMT state E (DESeq2 + apeglm)\n(* = |log2FC| >= 1 & padj <= 0.05)",
  row_title         = "BH4 Set Component Genes",
  cell_fun          = function(j, i, x, y, width, height, fill) {
    lfc  <- bh4_lfc_mat[i, j]
    padj <- bh4_padj_mat[i, j]
    if (!is.na(lfc) && !is.na(padj) && abs(lfc) >= 1 && padj <= 0.05) {
      grid::grid.text("*", x, y, gp = grid::gpar(fontsize = 16, fontface = "bold"))
    }
  }
)

svg("figures/ExtData5f_heatmap_bh4genes_lfc_sig.svg", width = 7, height = 6)
ComplexHeatmap::draw(bh4_lfc_heatmap_sig)
dev.off()

bh4_lfc_heatmap_sig
```

# Build a log2-transformed expression matrix for scoring

The scoring methods below (KS, GSVA, ssGSEA, UCell) all expect a numeric
gene x sample matrix on a roughly log scale, the same as
`TCGA_BRCA_EMT_GeneSet_Scoring.Rmd`'s already-log2 `data` layer. `protCoding`
is TPM, so log2(TPM + 1) puts it on a comparable scale.

```{r build-expr-matrix}
exprMat <- protCoding %>%
  tibble::column_to_rownames("Symbol") %>%
  as.matrix()

exprMat_log <- log2(exprMat + 1)

dim(exprMat_log)
```

# EMT scoring

## KS method (Chakraborty/Bell)

```{r source-ks-method}
source("../emt_code/Chakraborty_et_al_KS_code_vCAW.R")
```

```{r rescore-ks}
emt_genes <- unique(c(Epi.tum, Mes.tum))

expr_matrix_ks <- exprMat_log[rownames(exprMat_log) %in% emt_genes, ]
dim(expr_matrix_ks)

ks_scores <- suppressWarnings(KSScore(expr_matrix_ks, "SUM149PT", "."))
```

# Gene set scoring (GSVA / ssGSEA)

## GSVA

```{r score-gsva}
gsva_param  <- gsvaParam(exprMat_log, pathways_to_score, kcdf = "none")
gsva_scores <- gsva(gsva_param)
dim(gsva_scores)  # expect 9 gene sets x ncol(exprMat_log)

gsva_scores_df <- as.data.frame(t(gsva_scores))
colnames(gsva_scores_df) <- paste0(colnames(gsva_scores_df), "_GSVA")
```

## ssGSEA

Reuses `exprMat_log` and `pathways_to_score` from the GSVA step above,
scored with `ssgseaParam()` instead of `gsvaParam()`.

```{r score-ssgsea}
ssgsea_param  <- ssgseaParam(exprMat_log, pathways_to_score, normalize = TRUE)
ssgsea_scores <- gsva(ssgsea_param)
dim(ssgsea_scores)  # expect 9 gene sets x ncol(exprMat_log)

ssgsea_scores_df <- as.data.frame(t(ssgsea_scores))
colnames(ssgsea_scores_df) <- paste0(colnames(ssgsea_scores_df), "_ssGSEA")
```

## UCell

A 3rd scoring method, alongside GSVA and ssGSEA above: `UCell::AddModuleScore_UCell()`,
run on a `Seurat` object built from the same protein-coding, `Parental`-excluded
raw counts used for DESeq2 (`deseqCountMat`) -- UCell ranks genes per sample
from raw/normalized counts rather than working off `exprMat_log`. Same
`pathways_to_score` gene sets as GSVA/ssGSEA.

```{r score-ucell}
# UCell's default maxRank (1500) is tuned for sparse single-cell data, where
# most genes in a given cell are zero and only the top ~1500 ranked genes
# carry signal. Bulk RNA-seq samples are far denser (most protein-coding
# genes are detected at some level), so a maxRank that low would truncate
# the ranking before genes further down the list -- some of which may still
# be pathway members -- ever get ranked. Set much higher here to suit bulk
# data; tune this to see its effect on the resulting scores.
ucell_maxRank <- 10000

seurat_bulk <- Seurat::CreateSeuratObject(counts = deseqCountMat)

seurat_bulk <- UCell::AddModuleScore_UCell(
  seurat_bulk,
  features = pathways_to_score,
  maxRank  = ucell_maxRank
)

ucell_scores_df <- seurat_bulk@meta.data[, paste0(names(pathways_to_score), "_UCell"), drop = FALSE]
dim(ucell_scores_df)  # expect ncol(exprMat_log) samples x 9 gene sets
```

# Assemble per-sample score table

`sampleScores` plays the role `pseudobulk@meta.data` plays in the atlas
notebooks -- one row per sample (GSM), holding the design annotation plus
every computed score. `EMT_State` is ordered E -> EM1 -> EM2 -> EM3 -> M1
-> M2 (the induced epithelial-to-mesenchymal spectrum, `Parental` excluded
per above), per the experimental design.

```{r assemble-sample-scores}
sampleScores <- emt.state %>%
  dplyr::rename(GSM = GEO_Accession) %>%
  dplyr::filter(GSM %in% colnames(exprMat_log)) %>%
  dplyr::mutate(
    EMT_State = factor(EMT_State,
                        levels = c("E", "EM1", "EM2", "EM3", "M1", "M2")),
    EMT_KS_TumorSig = ks_scores[GSM]
  ) %>%
  dplyr::bind_cols(gsva_scores_df[.$GSM, , drop = FALSE]) %>%
  dplyr::bind_cols(ssgsea_scores_df[.$GSM, , drop = FALSE]) %>%
  dplyr::bind_cols(ucell_scores_df[.$GSM, , drop = FALSE])

table(sampleScores$EMT_State)
```

# Compare EMT scores across methods

```{r score-correlation-matrix}
score_cols <- c("EMT_KS_TumorSig", "Epi.tum_GSVA", "Mes.tum_GSVA")
cor(sampleScores[, score_cols], use = "pairwise.complete.obs")
```

# Compare GSVA vs ssGSEA scoring

Same 9 gene sets (`pathways_to_score`), GSVA score vs ssGSEA score for
each, one panel per gene set.

```{r gsva-vs-ssgsea-data}
gsva_ssgsea_long <- purrr::map_dfr(names(pathways_to_score), function(gs) {
  tibble(
    GeneSet = gs,
    GSVA    = sampleScores[[paste0(gs, "_GSVA")]],
    ssGSEA  = sampleScores[[paste0(gs, "_ssGSEA")]]
  )
}) %>%
  mutate(GeneSet = factor(GeneSet, levels = names(pathways_to_score)))
```

```{r gsva-vs-ssgsea-scatter, fig.width=10, fig.height=10}
ggplot(gsva_ssgsea_long, aes(x = GSVA, y = ssGSEA)) +
  geom_point(size = 1.2, alpha = 0.6) +
  geom_smooth(method = "lm", color = "red", linewidth = 0.6, se = FALSE) +
  ggpubr::stat_cor(size = 3) +
  facet_wrap(~ GeneSet, ncol = 3, scales = "free") +
  theme_bw() +
  labs(x = "GSVA score", y = "ssGSEA score",
       title = "GSVA vs ssGSEA score, by gene set (SUM149PT)")
```

# BH4 (GSVA) analysis

## BH4 gene expression vs EMT KS score

Raw (log2) expression for each of BH4's 7 genes, individually, against the
EMT KS score.

```{r bh4-genes-vs-emt-data}
bh4_gene_expr_long <- purrr::map_dfr(BH4, function(g) {
  tibble(GSM = colnames(exprMat_log), Gene = g, Expression = exprMat_log[g, ])
}) %>%
  dplyr::left_join(
    sampleScores %>% dplyr::select(GSM, EMT_KS_TumorSig),
    by = "GSM"
  )
```

```{r bh4-genes-vs-ks-scatter, fig.width=8, fig.height=12}
ggplot(bh4_gene_expr_long, aes(x = EMT_KS_TumorSig, y = Expression)) +
  geom_point(size = 1.2, alpha = 0.6) +
  geom_smooth(method = "lm", color = "red", linewidth = 0.6, se = FALSE) +
  ggpubr::stat_cor(size = 3) +
  facet_wrap(~ Gene, ncol = 2, scales = "free_y") +
  theme_bw() +
  labs(x = "EMT KS score", y = "Log2(TPM + 1)",
       title = "BH4 gene expression vs EMT KS score")
```

# Gene-set score heatmap

Gene-set-score rows (`Epi.tum`/`Mes.tum` right below the color bar, then
the pathway gene sets, in their existing order), samples ordered by
`EMT_State`, z-scored per row, with `EMT_State` as a color bar. No
individual genes are plotted. The GSVA heatmap keeps all 7 gene sets
(`Epi.tum`/`Mes.tum` + the 5 pathway sets); the ssGSEA heatmap below it
drops `Epi.tum`/`Mes.tum` and shows only the 5 pathway sets.

```{r genesetscore-heatmap-fn}
.geneSetScoreHeatmap <- function(scoreDf, exprMat, suffix, title,
                                   score_set_order = c("Epi.tum", "Mes.tum", "BH4", "Glutathione", "CoQ10", "Thioredoxin", "Peroxiredoxin")) {
  score_cols_hm <- paste0(score_set_order, suffix)

  sample_order <- order(scoreDf$EMT_State, scoreDf$GSM)
  ordered_df   <- scoreDf[sample_order, ]

  score_mat <- t(as.matrix(ordered_df[, score_cols_hm]))
  rownames(score_mat) <- sub(suffix, "", rownames(score_mat))

  combined_scaled <- t(scale(t(score_mat)))

  state_cols <- setNames(
    scales::hue_pal()(nlevels(scoreDf$EMT_State)),
    levels(scoreDf$EMT_State)
  )

  col_anno <- HeatmapAnnotation(
    EMT_State = ordered_df$EMT_State,
    col = list(EMT_State = state_cols)
  )

  Heatmap(combined_scaled,
          name              = "z-score",
          top_annotation    = col_anno,
          cluster_columns   = FALSE,
          cluster_rows      = FALSE,
          show_column_names = FALSE,
          column_title      = paste0(title, ", ordered by EMT state"),
          row_title         = "Gene set score")
}
```

```{r heatmap-gsva, fig.width=7, fig.height=4}
.geneSetScoreHeatmap(sampleScores, exprMat_log, "_GSVA", "SUM149PT (GSVA)")
```

```{r heatmap-ssgsea, fig.width=7, fig.height=4}
heatmap.ssgsea <- .geneSetScoreHeatmap(
  sampleScores, exprMat_log, "_ssGSEA", "SUM149PT (ssGSEA)",
  score_set_order = c("BH4", "Glutathione", "CoQ10", "Thioredoxin", "Peroxiredoxin")
)

# ggsave() only supports ggplot objects; a ComplexHeatmap needs a graphics
# device + draw() instead
# png("figures/heatmap_ssgsea.png", width = 11, height = 8, units = "in", res = 300)
# ComplexHeatmap::draw(heatmap.ssgsea)
# dev.off()

heatmap.ssgsea
```

# BH4 gene expression heatmap

Separate from the gene-set-score heatmap above: raw (log2) expression of
BH4's 7 member genes (gene-level values, not a gene-set score), samples
ordered by `EMT_State`, z-scored per row, with `EMT_State` as a color bar.
`Epi.tum`/`Mes.tum` gene-set scores are dropped from this view -- pass a
non-empty `score_set_order` to add them back (with a row-side strip
distinguishing score rows from gene rows) for another suffix/title.

```{r bh4gene-heatmap-fn}
.bh4GeneHeatmap <- function(scoreDf, exprMat, suffix, title, score_set_order = character(0)) {
  sample_order <- order(scoreDf$EMT_State, scoreDf$GSM)
  ordered_df   <- scoreDf[sample_order, ]

  gene_mat <- exprMat[BH4, ordered_df$GSM, drop = FALSE]

  if (length(score_set_order) > 0) {
    score_cols_hm <- paste0(score_set_order, suffix)
    score_mat     <- t(as.matrix(ordered_df[, score_cols_hm]))
    rownames(score_mat) <- sub(suffix, "", rownames(score_mat))
    combined_mat <- rbind(score_mat, gene_mat)
  } else {
    combined_mat <- gene_mat
  }

  combined_scaled <- t(scale(t(combined_mat)))

  state_cols <- setNames(
    scales::hue_pal()(nlevels(scoreDf$EMT_State)),
    levels(scoreDf$EMT_State)
  )

  col_anno <- HeatmapAnnotation(
    EMT_State = ordered_df$EMT_State,
    col = list(EMT_State = state_cols)
  )

  row_anno <- NULL
  row_title <- "BH4 Set Component Genes"
  if (length(score_set_order) > 0) {
    row_type <- factor(
      c(rep("Gene set score", length(score_set_order)), rep("Gene (BH4)", length(BH4))),
      levels = c("Gene set score", "Gene (BH4)")
    )
    row_anno <- rowAnnotation(
      `Row type` = row_type,
      col = list(`Row type` = c("Gene set score" = "#666666", "Gene (BH4)" = "#F2A104")),
      show_annotation_name = FALSE
    )
    row_title <- "Epi/Mes score / BH4 Set Component Genes"
  }

  Heatmap(combined_scaled,
          name              = "z-score",
          top_annotation    = col_anno,
          left_annotation   = row_anno,
          cluster_columns   = FALSE,
          cluster_rows      = FALSE,
          show_column_names = FALSE,
          column_title      = paste0(title, ", ordered by EMT state"),
          row_title         = row_title)
}
```

```{r bh4gene-heatmap, fig.width=7, fig.height=4}
bh4gene_heatmap <- .bh4GeneHeatmap(sampleScores, exprMat_log, "_ssGSEA", "SUM149PT BH4 genes")

# png("figures/heatmap_bh4genes.png", width = 7, height = 4, units = "in", res = 300)
# ComplexHeatmap::draw(bh4gene_heatmap)
# dev.off()

bh4gene_heatmap
```

Same layout as above (samples ordered by `EMT_State`, same column annotation),
but showing normalized log2 expression values directly instead of z-scoring
each row.

```{r bh4gene-heatmap-log2-fn}
.bh4GeneHeatmapLog2 <- function(scoreDf, exprMat, suffix, title, score_set_order = character(0)) {
  sample_order <- order(scoreDf$EMT_State, scoreDf$GSM)
  ordered_df   <- scoreDf[sample_order, ]

  gene_mat <- exprMat[BH4, ordered_df$GSM, drop = FALSE]

  if (length(score_set_order) > 0) {
    score_cols_hm <- paste0(score_set_order, suffix)
    score_mat     <- t(as.matrix(ordered_df[, score_cols_hm]))
    rownames(score_mat) <- sub(suffix, "", rownames(score_mat))
    combined_mat <- rbind(score_mat, gene_mat)
  } else {
    combined_mat <- gene_mat
  }

  state_cols <- setNames(
    scales::hue_pal()(nlevels(scoreDf$EMT_State)),
    levels(scoreDf$EMT_State)
  )

  col_anno <- HeatmapAnnotation(
    EMT_State = ordered_df$EMT_State,
    col = list(EMT_State = state_cols)
  )

  row_anno <- NULL
  row_title <- "BH4 Set Component Genes"
  if (length(score_set_order) > 0) {
    row_type <- factor(
      c(rep("Gene set score", length(score_set_order)), rep("Gene (BH4)", length(BH4))),
      levels = c("Gene set score", "Gene (BH4)")
    )
    row_anno <- rowAnnotation(
      `Row type` = row_type,
      col = list(`Row type` = c("Gene set score" = "#666666", "Gene (BH4)" = "#F2A104")),
      show_annotation_name = FALSE
    )
    row_title <- "Epi/Mes score / BH4 Set Component Genes"
  }

  Heatmap(combined_mat,
          name              = "log2 expression",
          top_annotation    = col_anno,
          left_annotation   = row_anno,
          cluster_columns   = FALSE,
          cluster_rows      = FALSE,
          show_column_names = FALSE,
          column_title      = paste0(title, ", ordered by EMT state"),
          row_title         = row_title)
}
```

```{r bh4gene-heatmap-log2, fig.width=7, fig.height=4}
bh4gene_heatmap_log2 <- .bh4GeneHeatmapLog2(sampleScores, exprMat_log, "_ssGSEA", "SUM149PT BH4 genes")

# png("figures/heatmap_bh4genes_log2.png", width = 7, height = 4, units = "in", res = 300)
# ComplexHeatmap::draw(bh4gene_heatmap_log2)
# dev.off()

bh4gene_heatmap_log2
```

# BH4 gene expression bar charts by replicate

One panel per BH4 gene, bars grouped by `EMT_State`, one bar per replicate
within each state (samples ordered by `GSM` within `EMT_State`, same
grouping used for the heatmaps above). Two versions: per-gene z-scored
values, and raw normalized log2 expression values.

```{r bh4-gene-barchart-data}
bh4_bar_df <- purrr::map_dfr(BH4, function(g) {
  tibble(GSM = colnames(exprMat_log), Gene = g, Log2Expression = exprMat_log[g, ])
}) %>%
  dplyr::left_join(sampleScores %>% dplyr::select(GSM, EMT_State), by = "GSM") %>%
  dplyr::group_by(Gene) %>%
  dplyr::mutate(ZScore = as.numeric(scale(Log2Expression))) %>%
  dplyr::group_by(Gene, EMT_State) %>%
  dplyr::arrange(GSM, .by_group = TRUE) %>%
  dplyr::mutate(Replicate = factor(paste0("Rep", dplyr::row_number()))) %>%
  dplyr::ungroup()
```

```{r bh4-gene-barchart-fn}
.bh4GeneBarChart <- function(df, value_col, y_lab, title) {
  pdf <- df %>% dplyr::rename(Value = all_of(value_col))

  ggplot(pdf, aes(x = EMT_State, y = Value, fill = Replicate)) +
    geom_col(position = position_dodge(width = 0.8), width = 0.7,
             color = "black", linewidth = 0.3) +
    facet_wrap(~ Gene, ncol = 4, scales = "free_y") +
    theme_bw() +
    theme(
      panel.grid   = element_blank(),
      panel.border = element_rect(colour = "black", fill = NA, linewidth = 1.2),
      axis.title   = element_text(face = "bold", size = 13),
      axis.text    = element_text(face = "bold", size = 10)
    ) +
    labs(x = "EMT state", y = y_lab, fill = "Replicate", title = title)
}
```

```{r bh4-gene-barchart-zscore, fig.width=12, fig.height=8}
.bh4GeneBarChart(bh4_bar_df, "ZScore", "Z-score",
                  "SUM149PT BH4 gene expression (z-score) by EMT state")
```

```{r bh4-gene-barchart-log2, fig.width=12, fig.height=8}
.bh4GeneBarChart(bh4_bar_df, "Log2Expression", "Log2(TPM + 1)",
                  "SUM149PT BH4 gene expression (log2) by EMT state")
```

# Single-panel gene set boxplots by EMT state

One panel per gene set/method, box + jitter by `EMT_State`, tight
whisker-based y-axis (`.tightBoxYlim()`, defined below). No significance
brackets here, given the same n=3-per-group density concern as noted for
`.pairwiseTests()` below.

```{r tightboxylim-fn}
.tightBoxYlim <- function(df, y_col, group_col) {
  group_ranges <- tapply(df[[y_col]], df[[group_col]], function(y) {
    range(boxplot.stats(y)$stats)
  })
  c(min(sapply(group_ranges, min)), max(sapply(group_ranges, max)))
}

# All pairwise Welch's t-tests among group_col's levels, BH-adjusted --
# used below for the BH4-with-stats boxplot; not invoked on the denser
# 9-gene-set x 7-level EMT_State panels above (too many comparisons to
# read at n=3 per group). Welch's t-test, not Wilcoxon rank-sum -- at
# n=3-per-group, the exact Wilcoxon test's null distribution is so coarse
# (only choose(6,3)=20 possible splits) that its minimum achievable
# two-sided p-value is 0.1, i.e. p < 0.05 is mathematically unreachable
# regardless of effect size. `ref_group`, when given, restricts the pairs
# tested (and therefore the BH adjustment) to only those involving that
# group -- e.g. every level vs "E".
.pairwiseTests <- function(df, y_col, group_col, min_n = 2, ref_group = NULL) {
  groups <- levels(df[[group_col]])
  pairs  <- utils::combn(groups, 2, simplify = FALSE)

  if (!is.null(ref_group)) {
    pairs <- purrr::keep(pairs, function(pr) ref_group %in% pr)
  }

  res <- purrr::map_dfr(pairs, function(pr) {
    x <- df[[y_col]][df[[group_col]] == pr[1]]
    y <- df[[y_col]][df[[group_col]] == pr[2]]
    if (length(x) < min_n || length(y) < min_n) return(tibble())
    tibble(group1 = pr[1], group2 = pr[2], p = t.test(x, y)$p.value)
  })

  if (nrow(res) == 0) return(res)

  res$p.adj       <- p.adjust(res$p, method = "BH")
  res$significant <- res$p.adj < 0.05
  res$p.format    <- formatC(res$p.adj, format = "g", digits = 3)
  res
}
```

```{r genesetscore-single-panel-fn}
.geneSetSinglePanel <- function(scoreDf, gene_set, group_var, x_lab, method = c("GSVA", "ssGSEA", "UCell")) {
  method    <- match.arg(method)
  score_col <- paste0(gene_set, "_", method)

  pdf <- scoreDf %>% dplyr::select(Score = all_of(score_col), Group = all_of(group_var))
  ylim_vals <- .tightBoxYlim(pdf, "Score", "Group")
  group_n   <- table(pdf$Group)

  ggplot(pdf, aes(x = Group, y = Score, fill = Group)) +
    geom_boxplot(outlier.shape = NA, alpha = 0.8, linewidth = 0.7) +
    geom_jitter(width = 0.15, alpha = 0.6, size = 1) +
    coord_cartesian(ylim = ylim_vals) +
    theme_bw() +
    theme(
      panel.grid      = element_blank(),
      panel.border    = element_rect(colour = "black", fill = NA, linewidth = 1.2),
      axis.title      = element_text(face = "bold", size = 13),
      axis.text       = element_text(face = "bold", size = 11),
      legend.position = "none"
    ) +
    scale_x_discrete(labels = function(l) paste0(l, "\n(n=", group_n[l], ")")) +
    labs(x = NULL, y = paste0(gene_set, " score (", method, ")"),
         title = paste0(gene_set, " ", method, " score by ", x_lab, " (SUM149PT)"))
}
```

```{r genesets-ssgsea-boxplot-by-emtstate, fig.width=6, fig.height=4}
for (gs in names(pathways_to_score)) {
  print(.geneSetSinglePanel(sampleScores, gs, "EMT_State", "EMT state", method = "ssGSEA"))
}
```

## BH4 ssGSEA score by EMT state (saved)

Single-panel `BH4` ssGSEA figure from the loop above, saved on its own.

```{r bh4-ssgsea-boxplot-by-emtstate, fig.width=5, fig.height=5}
bh4_ssgsea_boxplot <- .geneSetSinglePanel(sampleScores, "BH4", "EMT_State", "EMT state", method = "ssGSEA")

# ggsave("figures/bh4_ssgsea_boxplot_by_emtstate.png", plot = bh4_ssgsea_boxplot, width = 6, height = 4, dpi = 300)

bh4_ssgsea_boxplot
```

## UCell boxplots by EMT state

Same style as the ssGSEA panels above, one per gene set, now using the
UCell scores (`ucell_maxRank` above).

```{r genesets-ucell-boxplot-by-emtstate, fig.width=6, fig.height=4}
for (gs in names(pathways_to_score)) {
  print(.geneSetSinglePanel(sampleScores, gs, "EMT_State", "EMT state", method = "UCell"))
}
```

## BH4 UCell score by EMT state (saved)

Single-panel `BH4` UCell figure from the loop above, saved on its own.

```{r bh4-ucell-boxplot-by-emtstate, fig.width=5, fig.height=5}
bh4_ucell_boxplot <- .geneSetSinglePanel(sampleScores, "BH4", "EMT_State", "EMT state", method = "UCell")

# ggsave("figures/bh4_ucell_boxplot_by_emtstate.png", plot = bh4_ucell_boxplot, width = 6, height = 4, dpi = 300)

bh4_ucell_boxplot
```

## BH4 score by EMT state, with pairwise statistics

Alternate version of the BH4 boxplot with BH-adjusted pairwise Welch's
t-test brackets (`.pairwiseTests()` above) drawn in -- restricted to each
remaining level vs `E` only (all other pairwise comparisons are dropped),
so the BH adjustment itself is computed over just those comparisons.
Brackets are stacked shortest-span-first so nested comparisons stay
readable-ish. One version per scoring method (GSVA, ssGSEA, UCell).

```{r bh4-stats-boxplot-fn}
.bh4StatsBoxplot <- function(scoreDf, method) {
  score_col    <- paste0("BH4_", method)
  bh4_state_df <- scoreDf %>% dplyr::select(Score = all_of(score_col), Group = EMT_State)

  bh4_pairwise_stats <- .pairwiseTests(bh4_state_df, "Score", "Group", ref_group = "E")

  group_levels <- levels(bh4_state_df$Group)
  bh4_pairwise_stats <- bh4_pairwise_stats %>%
    dplyr::mutate(
      xmin_pos = match(group1, group_levels),
      xmax_pos = match(group2, group_levels),
      span     = xmax_pos - xmin_pos
    ) %>%
    dplyr::arrange(span, xmin_pos)

  data_max  <- max(bh4_state_df$Score, na.rm = TRUE)
  data_step <- diff(range(bh4_state_df$Score, na.rm = TRUE)) * 0.06
  bh4_pairwise_stats$y.position <- data_max + data_step * seq_len(nrow(bh4_pairwise_stats))

  group_n <- table(bh4_state_df$Group)

  ggplot(bh4_state_df, aes(x = Group, y = Score, fill = Group)) +
    geom_boxplot(outlier.shape = NA, alpha = 0.8, linewidth = 0.7) +
    geom_jitter(width = 0.15, alpha = 0.6, size = 1) +
    theme_bw() +
    theme(
      panel.grid      = element_blank(),
      panel.border    = element_rect(colour = "black", fill = NA, linewidth = 1.2),
      axis.title      = element_text(face = "bold", size = 13),
      axis.text       = element_text(face = "bold", size = 11),
      legend.position = "none"
    ) +
    scale_x_discrete(labels = function(l) paste0(l, "\n(n=", group_n[l], ")")) +
    labs(x = NULL, y = paste0("BH4 score (", method, ")"),
         title = paste0("BH4 score by EMT state (SUM149PT, ", method, ")")) +
    ggpubr::stat_pvalue_manual(bh4_pairwise_stats, label = "p.format",
                                xmin = "group1", xmax = "group2",
                                y.position = "y.position",
                                tip.length = 0.005, bracket.size = 0.3,
                                size = 4, fontface = "bold")
}
```

```{r bh4-boxplot-by-emtstate-stats, fig.width=6, fig.height=5}
for (m in c("GSVA", "ssGSEA", "UCell")) {
  bh4_boxplot_stats <- .bh4StatsBoxplot(sampleScores, m)

  # ggsave(paste0("figures/bh4_boxplot_by_emtstate_stats_", m, ".png"),
  #        plot = bh4_boxplot_stats, width = 6, height = 5, dpi = 300)

  print(bh4_boxplot_stats)
}
```

Same BH4-by-`EMT_State` comparison, but with an ordinary one-way ANOVA
(`aov()`) followed by Dunnett's multiple-comparisons test
(`multcomp::glht()`, `mcp(Group = "Dunnett")`) in place of the BH-adjusted
pairwise Welch's t-tests above, with `E` set as the reference level.
Dunnett's test is built for exactly this "every group vs. one reference"
design -- it accounts for the correlation among comparisons that share a
common reference group, so its adjustment is less conservative here than
BH across independent pairwise tests. It still assumes normal,
equal-variance residuals across groups (the ANOVA omnibus test), which is
a stretch at n=3/group, so treat this as a second opinion rather than a
replacement. Again, one version per scoring method.

```{r bh4-dunnett-boxplot-fn}
.bh4DunnettBoxplot <- function(scoreDf, method) {
  score_col    <- paste0("BH4_", method)
  bh4_state_df <- scoreDf %>% dplyr::select(Score = all_of(score_col), Group = EMT_State)

  bh4_dunnett_df <- bh4_state_df %>%
    dplyr::mutate(Group = stats::relevel(Group, ref = "E"))

  bh4_aov <- aov(Score ~ Group, data = bh4_dunnett_df)
  print(summary(bh4_aov))

  bh4_dunnett     <- multcomp::glht(bh4_aov, linfct = multcomp::mcp(Group = "Dunnett"))
  dunnett_summary <- summary(bh4_dunnett)
  print(dunnett_summary)

  bh4_dunnett_stats <- tibble::tibble(
    group1 = "E",
    group2 = sub(" - E$", "", names(dunnett_summary$test$coefficients)),
    p.adj  = as.numeric(dunnett_summary$test$pvalues)
  ) %>%
    dplyr::mutate(
      significant = p.adj < 0.05,
      p.format    = formatC(p.adj, format = "g", digits = 3)
    )

  group_levels <- levels(bh4_state_df$Group)
  bh4_dunnett_stats <- bh4_dunnett_stats %>%
    dplyr::mutate(
      xmin_pos = match(group1, group_levels),
      xmax_pos = match(group2, group_levels),
      span     = xmax_pos - xmin_pos
    ) %>%
    dplyr::arrange(span, xmin_pos)

  data_max  <- max(bh4_state_df$Score, na.rm = TRUE)
  data_step <- diff(range(bh4_state_df$Score, na.rm = TRUE)) * 0.06
  bh4_dunnett_stats$y.position <- data_max + data_step * seq_len(nrow(bh4_dunnett_stats))

  group_n <- table(bh4_state_df$Group)

  ggplot(bh4_state_df, aes(x = Group, y = Score, fill = Group)) +
    geom_boxplot(outlier.shape = NA, alpha = 0.8, linewidth = 0.7) +
    geom_jitter(width = 0.15, alpha = 0.6, size = 1) +
    theme_bw() +
    theme(
      panel.grid      = element_blank(),
      panel.border    = element_rect(colour = "black", fill = NA, linewidth = 1.2),
      axis.title      = element_text(face = "bold", size = 13),
      axis.text       = element_text(face = "bold", size = 11),
      legend.position = "none"
    ) +
    scale_x_discrete(labels = function(l) paste0(l, "\n(n=", group_n[l], ")")) +
    labs(x = NULL, y = paste0("BH4 score (", method, ")"),
         title = paste0("BH4 score by EMT state (SUM149PT, ", method, ")\nANOVA + Dunnett's vs. E")) +
    ggpubr::stat_pvalue_manual(bh4_dunnett_stats, label = "p.format",
                                xmin = "group1", xmax = "group2",
                                y.position = "y.position",
                                tip.length = 0.005, bracket.size = 0.3,
                                size = 4, fontface = "bold")
}
```

```{r bh4-boxplot-by-emtstate-dunnett, fig.width=6, fig.height=5}
# for (m in c("GSVA", "ssGSEA", "UCell")) {
for (m in c("ssGSEA")) {
  bh4_boxplot_dunnett <- .bh4DunnettBoxplot(sampleScores, m)

  ggsave(paste0("figures/ExtData5e_bh4_boxplot_by_emtstate_dunnett.svg"),
         plot = bh4_boxplot_dunnett, width = 6, height = 5, dpi = 300)

  print(bh4_boxplot_dunnett)
}
```

# Manual dot plots

Mean expression / pct-expressed dot plot, grouped by `EMT_State`. `pctExp`
(% samples with expression > 0 within a group) is a much weaker signal for
bulk TPM data than for single-cell dropout-prone data -- it will often sit
near 100% -- so `normExp` (min-max normalized mean expression) is the more
informative color channel here.

```{r manual-dotplot-fn}
.manualDotPlotData <- function(exprMat, scoreDf, genes, group_var) {
  expr_sub     <- exprMat[genes, scoreDf$GSM, drop = FALSE]
  group_levels <- levels(scoreDf[[group_var]])

  purrr::map_dfr(group_levels, function(g) {
    samples_in_group <- scoreDf$GSM[scoreDf[[group_var]] == g]
    expr_g <- expr_sub[, samples_in_group, drop = FALSE]
    tibble(
      gene    = genes,
      group   = g,
      meanExp = rowMeans(expr_g),
      pctExp  = rowMeans(expr_g > 0) * 100
    )
  }) %>%
    dplyr::group_by(gene) %>%
    dplyr::mutate(
      zscore  = as.numeric(scale(meanExp)),
      normExp = (meanExp - min(meanExp)) / (max(meanExp) - min(meanExp))
    ) %>%
    dplyr::ungroup() %>%
    dplyr::mutate(
      gene  = factor(gene, levels = genes),
      group = factor(group, levels = group_levels)
    )
}

.manualDotPlot <- function(dot_long, x_lab = "Gene", y_lab = "Group",
                            color_by = c("zscore", "normExp")) {
  color_by <- match.arg(color_by)

  if (color_by == "zscore") {
    lim <- max(abs(dot_long$zscore), na.rm = TRUE)
    lim <- ifelse(lim == 0, 1, lim)

    p <- ggplot(dot_long, aes(x = gene, y = group)) +
      geom_point(aes(size = pctExp, color = zscore)) +
      scale_color_gradientn(
        colors = c("#2F6FB2", "#9FC3DF", "#F4EFE6", "#F28E5C", "#E84A3A"),
        values = scales::rescale(c(-lim, -lim / 3, 0, lim / 3, lim)),
        limits = c(-lim, lim),
        name   = "Z-score\nmean expr."
      )
  } else {
    p <- ggplot(dot_long, aes(x = gene, y = group)) +
      geom_point(aes(size = pctExp, color = normExp)) +
      scale_color_gradientn(
        colors = c("#2F6FB2", "#9FC3DF", "#F4EFE6", "#F28E5C", "#E84A3A"),
        limits = c(0, 1),
        name   = "Normalized\nmean expr."
      )
  }

  p +
    scale_size_continuous(name = "% expressed", range = c(0.5, 10)) +
    theme_minimal(base_size = 10) +
    theme(
      axis.text.x  = element_text(angle = 45, hjust = 1, face = "italic"),
      panel.grid   = element_blank(),
      panel.border = element_rect(colour = "grey40", fill = NA, linewidth = 0.6)
    ) +
    labs(x = x_lab, y = y_lab)
}
```

## BH4 genes: dot plot by EMT state

```{r dotplot-bh4-by-emtstate, fig.width=8, fig.height=6}
bh4_dot_state <- .manualDotPlotData(exprMat_log, sampleScores, BH4, "EMT_State")
.manualDotPlot(bh4_dot_state, x_lab = "Gene", y_lab = "EMT state", color_by = "normExp")
```

# write session info

Capturing information about the R session helps document your work

```{r session-info}
sessionInfo()
writeLines(capture.output(sessionInfo()), "SUM149PT_EMT_scoring_plotting_sessionInfo.txt")
```
