---
title: "TCGA PANCAN RNAseq - Data Import and preparation"
author: "Charlie Whittaker"
output:
  html_document:
    toc: true
    toc_depth: 3
  html_notebook: default
date: "8/3/2026"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(future.globals.maxSize = 5e10)
set.seed(1234)
```

# Goal

Load the TCGA PANCAN batch-corrected expression matrix and the matched
clinical/metadata table, assemble a single Seurat object (one "cell" per
tumor sample, following the convention used elsewhere in this project), and
save it for downstream EMT/gene-set scoring.

## Raw data

`data/geneexpressionmatrix_uniqueIDs.txt` (the Firehose `EBPlusPlusAdjustPANCAN`
batch-corrected matrix) and `data/Updated_Metadata_fixed.txt` (matched clinical
metadata) are from the TCGA Pan-Cancer Atlas project. The GDC/Broad Firehose
pages that originally hosted these files are no longer live, so no working
download link is available; the traceable provenance reference for this
dataset is the associated publication:

Liu, J., Lichtenberg, T., Hoadley, K.A. et al. An Integrated TCGA Pan-Cancer
Clinical Data Resource to Drive High-Quality Survival Outcome Analytics.
*Cell* 173(2), 400–416.e11 (2018).
<https://doi.org/10.1016/j.cell.2018.02.052>
(<https://www.sciencedirect.com/science/article/pii/S0092867418302290>)

```{r load-packages, message=FALSE, warning=FALSE}
library(tidyverse)
library(Matrix)
library(Seurat)
```

# Load expression matrix

`geneexpressionmatrix_uniqueIDs.txt` is the Broad Firehose
`EBPlusPlusAdjustPANCAN` batch-corrected RSEM matrix (genes x samples,
continuous values -- not raw counts). `read.table()`'s default
`check.names = TRUE` runs `make.names()` on the sample-barcode column
headers, converting the hyphens in TCGA barcodes (e.g.
`TCGA-02-0047-01A-01R-1849-01`) to dots -- this is why the metadata barcodes
get the same substitution below, to keep the two joinable.

```{r load-expr-matrix}
gene_expr <- read.table("data/geneexpressionmatrix_uniqueIDs.txt", header = TRUE, row.names = 1, sep = "\t")
dim(gene_expr)
```

```{r clean-expr-matrix}
# Replace non-finite values (NA/NaN/Inf) with 0. Deliberately NOT rounding to
# integers here (unlike the earlier TCGA_analysis_*.Rmd scripts) -- these are
# already continuous batch-corrected values, and the EMT-scoring methods
# (KSScore, GSVA, UCell) all expect continuous expression, so rounding would
# only throw away precision for no benefit.
gene_expr <- gene_expr %>% mutate(across(everything(), ~replace(., is.na(.) | is.infinite(.) | is.nan(.), 0)))
```

# Build the Seurat object

```{r create-seurat-object}
seurat_obj <- CreateSeuratObject(gene_expr)
seurat_obj
```

# Attach sample metadata

```{r load-metadata}
metadata <- read.delim("data/Updated_Metadata_fixed.txt")
metadata$bcr_patient_barcode <- gsub("-", ".", metadata$bcr_patient_barcode)
```

```{r dedup-metadata}
nrow(metadata)
metadata <- distinct(metadata)
nrow(metadata)
```

Index by barcode and hand the whole table to `AddMetaData()` in one call --
this replaces the old pattern of one `setNames()` + `AddMetaData()` pair per
column (30 repetitions in the earlier scripts) with a single call that adds
every column at once; samples in `seurat_obj` without a metadata match get
`NA` for all fields, same as before.

```{r attach-metadata}
rownames(metadata) <- metadata$bcr_patient_barcode
seurat_obj <- AddMetaData(seurat_obj, metadata = metadata)

head(seurat_obj@meta.data)
```

# Sanity checks

```{r sanity-check-coverage}
# How many samples in the expression matrix got no metadata match?
sum(is.na(seurat_obj$type))

# Sample counts per cancer type
table(seurat_obj$type, useNA = "ifany")
```

# QC violin plots

```{r qc-violin, fig.width=8, fig.height=5}
VlnPlot(seurat_obj, features = c("nCount_RNA", "nFeature_RNA"))
```

type_order <- tcga@meta.data %>%
  group_by(type) %>%
  summarize(median_nCount = median(nCount_RNA, na.rm = TRUE)) %>%
  arrange(desc(median_nCount)) %>%
  pull(type)

tcga$type <- factor(tcga$type, levels = type_order)

VlnPlot(tcga, features = "nCount_RNA", group.by = "type", pt.size = 0)
```

# Subset PANCANCER object to BRCA

```{r load-and-subset}
tcga_brca <- subset(seurat_obj, subset = type == "BRCA")

ncol(seurat_obj)   # pan-cancer sample count
ncol(tcga_brca)  # BRCA-only sample count
```

# Save BRCA object

```{r}
saveRDS(tcga_brca, file="TCGA_BRCA_seurat.rds")
```

# write session info

Capturing information about the R session helps document your work

```{r session-info}
sessionInfo()
writeLines(capture.output(sessionInfo()), "TCGA_BRCA_prep_sessionInfo.txt")
```
