---
title: "Whitney Henry - Bulk RNAseq Analysis - SKCM TCGA (Tsoi 2018, PMID 29657129) - BH4 Pathway Scoring Figures"
author: "Evelyn Tong"
output:
  html_document:
    toc: true
    toc_depth: 3
  html_notebook: default
date: "9/22/2026"
---

Split out of the original monolithic `tmp/WH_BH4pathway_final/src/Analysis_script.Rmd`
("BH4 analysis script - Evelyn - 2026/09/16", Data4 section) during the
`datasets/` reorg, into its own self-contained dataset directory/project.
See `SKCM_TCGA_BH4_scoring_Methods.md` for the full methodology.

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, message = FALSE, warning = FALSE,
  fig.align = "center", fig.width = 6, fig.height = 4, dpi = 300
)
source("emt_code/bh4_pathway_shared_setup.R")
```

## parameter settings

```{r}
# figures/ + stats output for this dataset
current.outputfolder <- "figures"
# define where the input RDS file is
data.path <- "TCGA_bulkRNAseq_skcm.rds"

# define state categories and colors
STATE_ORDER <- c("Melanocytic", "Transitory", "Neural crest-like", "Undifferentiated")
state_color <- c("Melanocytic"       = "#4DAF4A",   # green  (C4)
                 "Transitory"        = "#984EA3",   # purple (C3)
                 "Neural crest-like" = "#E41A1C",   # red    (C2)
                 "Undifferentiated"  = "#377EB8")   # blue   (C1)

# ── Tissue Site -> point shape (second boxplot) ──
tissue_shapes <- c("Primary Tumor"                             = 16,
                   "Distant Metastasis"                        = 17,
                   "Regional Lymph Node"                       = 15,
                   "Regional Cutaneous or Subcutaneous Tissue" = 3)
```

## load data and processing

```{r}
seu <- readRDS(data.path)
seu$Patient_ID <- gsub("TCGA-SKCM_", "", rownames(seu@meta.data))

cat("  Dimensions:", dim(seu)[1], "genes x", dim(seu)[2], "cells\n")
cat("  BH4 genes present:", sum(bh4_genes %in% rownames(seu)), "/", length(bh4_genes), "\n")

seu <- NormalizeData(seu)
```

## define state based on 4-state trajectory from Tsoi 2018;PMID 29657129

```{r}
# read in the state definition from the paper (PMID: 29657129)
state_assignment <- read_excel_allsheets("SKCM_Tsoi2018_suppTable_s4.xlsx")$`Table S4`
colnames(state_assignment) <- state_assignment[1, ]
state_assignment <- state_assignment[-1, ]
state_assignment$Patient_ID <- sub("^([^-]+-[^-]+-[^-]+).*", "\\1", state_assignment$Sample)
state_assignment <- state_assignment %>% group_by(Patient_ID) %>% slice_head(n = 1) %>% as.data.frame()
rownames(state_assignment) <- paste0("TCGA-SKCM_", state_assignment$Patient_ID)

seu <- AddMetaData(seu, metadata = state_assignment)
```

## calculate BH4 signature score (ssgsea)

```{r ssgsea}
# Expression matrix (genes x samples). Using the `data` layer; ssGSEA is rank-based
# so this is invariant to the transform. If your object only has counts, that layer
# is used as-is (still fine for ssGSEA ranks).
expr <- as.matrix(GetAssayData(seu, assay = "RNA", layer = "data"))

# Modern GSVA param-object API (matches BRCA_TCGA/BRCA_atlas_CellxGene_Chen):
# GSVA >= 2.x removed the old gsva(matrix, sets, method = ...) call in favor
# of building a Param object and passing that to gsva().
ssgsea_param <- ssgseaParam(expr, bh4_sigs, normalize = TRUE)
ssg <- gsva(ssgsea_param)
ssg_t <- t(ssg)
for (s in bh4_score_names)
  seu@meta.data[[paste0(s, "_ssGSEA")]] <- ssg_t[rownames(seu@meta.data), s]
```

## plotting
### Main 5g boxplot by BH4 score
```{r}
# Abbreviated x-axis labels so the four subtype breaks don't overlap.
short_state <- function(v) {
  v[v == "Undifferentiated"]  <- "Undiff"
  v[v == "Neural crest-like"] <- "NCL"
  v[v == "Transitory"]        <- "Transit"
  v[v == "Melanocytic"]       <- "Melano"
  v
}
# Use the updated Seurat object name
md <- seu@meta.data

# Only plot pathway scores (no individual genes)
quantities <- bh4_score_names

# Build long dataframe for the pathways
box_long <- do.call(rbind, lapply(bh4_score_names, function(s) {
  data.frame(
    Quantity = s,
    cancer_state = md$Subtype,
    value = as.numeric(md[[paste0(s, "_ssGSEA")]]),
    y_label = "ssGSEA score",
    stringsAsFactors = FALSE
  )
}))

# Apply abbreviations directly to the dataframe
box_long$cancer_state <- short_state(box_long$cancer_state)
# Apply abbreviations to the order and fill vectors to ensure make_boxplot_state matches perfectly
short_levels <- short_state(STATE_ORDER)
short_fills <- state_color
names(short_fills) <- short_state(names(state_color))

# (1) standard boxplots (using previously defined make_boxplot_state)
make_boxplot_state(
  box_long,
  quantities,
  title_str = "BH4 pathways by Tsoi subtype  |  ssGSEA",
  out_pdf   = file.path(current.outputfolder, "5g_Boxplot_byCellState_BH4_score.pdf"),
  levels    = short_levels,
  fills     = short_fills
)
```

## write session info
```{r session-info}
sessionInfo()
writeLines(capture.output(sessionInfo()), "SKCM_TCGA_BH4_scoring_sessionInfo.txt")
```
