---
title: "Whitney Henry - scRNAseq Analysis - PDAC human (Park, PMID 38702773) - BH4 Pathway Scoring Figures"
author: "Evelyn Tong"
output:
  html_document:
    toc: true
    toc_depth: 3
  html_notebook: default
date: "9/22/2026"
---

Split out of the original monolithic `tmp/WH_BH4pathway_final/src/Analysis_script.Rmd`
("BH4 analysis script - Evelyn - 2026/09/16", Data2 section) during the
`datasets/` reorg, into its own self-contained dataset directory/project.
See `PDAC_human_park_BH4_scoring_Methods.md` for the full methodology.

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, message = FALSE, warning = FALSE,
  fig.align = "center", fig.width = 6, fig.height = 4, dpi = 300
)
source("emt_code/bh4_pathway_shared_setup.R")
```

## parameter settings
```{r}
# figures/ + stats output for this dataset
current.outputfolder <- "figures"
# define where the input RDS file is
data.path <- "park_bh4_seurat.rds"
# define state categories and colors
STATE_ORDER <- c("Classical", "IC", "Basal")
state_color <- c("Classical"     = "#334589",
                 "IC" = "#148241",
                 "Basal"       = "#ed9d2d")
```
## load data and processing
```{r}
seu <- readRDS(data.path) # This object is already subset to malignant cells (copykat), so the malignant compartment is the whole object -> median nFeature_RNA over all cells.
cat("  Dimensions:", dim(seu)[1], "genes x", dim(seu)[2], "cells\n")
cat("  BH4 genes present:", sum(bh4_genes %in% rownames(seu)), "/", length(bh4_genes), "\n")

seu <- NormalizeData(seu)

# ── Dataset-customized UCell maxRank ─────────────────────────────────────────
# Malignant compartment = cells used for plotting/correlation (non-"Other").
if (!"nFeature_RNA" %in% colnames(seu@meta.data)) {
  seu$nFeature_RNA <- Matrix::colSums(GetAssayData(seu, assay = "RNA", layer = "counts") > 0)
}
ucell_maxrank_custom <- as.integer(round(median(seu$nFeature_RNA)))

cat(sprintf(
  "\n  >>> Custom UCell maxRank = median nFeature_RNA of malignant compartment\n"))
cat(sprintf(
  "      median nFeature_RNA (chosen maxRank)   : %d\n", ucell_maxrank_custom))
```
## define state based on raghavan signature
```{r}
ic_threshold = 0.2
sc.signatures <- readxl::read_excel("34890551_Raghavan_state_signatures.xlsx")
sig_list <- list(
  scBasal     = na.omit(sc.signatures$scBasal[sc.signatures$scBasal != ""]),
  scClassical = na.omit(sc.signatures$scClassical[sc.signatures$scClassical != ""]),
  IC          = na.omit(sc.signatures$IC[sc.signatures$IC != ""])
)
for (nm in names(sig_list)) {
  seu <- AddModuleScore(
    seu,
    features = list(sig_list[[nm]]),
    name = nm,
    search = FALSE
  )
  colnames(seu@meta.data)[colnames(seu@meta.data) == paste0(nm, "1")] <- paste0(nm, "_AMS")
}

# 4. Classify sc-set state (Raghavan 2021 rule) using the AMS scores
clas <- seu@meta.data[["scClassical_AMS"]]
bas  <- seu@meta.data[["scBasal_AMS"]]
ic   <- seu@meta.data[["IC_AMS"]]

tt_cb    <- clas - bas          # Classical - Basal
ic_score <- ic - abs(tt_cb)     # IC_Score per the original paper

seu$cell_state <- dplyr::case_when(
  ic_score >= ic_threshold                 ~ "IC",
  ic_score <  ic_threshold & tt_cb > 0     ~ "Classical",
  ic_score <  ic_threshold & tt_cb < 0     ~ "Basal",
  TRUE                                     ~ "undefined"
)
seu$cell_state <- factor(as.character(seu@meta.data$cell_state), levels = STATE_ORDER)
cat("  State counts:\n"); print(table(seu$cell_state, useNA = "ifany"))
```

## calculate BH4 signature score
```{r}
# Run UCell, maxRank = dataset-customized maxRank (median nFeature of malignant)
seu <- AddModuleScore_UCell(
  seu,
  features = bh4_sigs,
  maxRank = ucell_maxrank_custom,
  name = "_UCellMedianMaxRank",
  ncores = 4
)

# Composite (full-pathway) scores for the summary / UMAP / heatmap
seu$BH4_UCellMedianMaxRank <- seu@meta.data[[paste0(bh4_full, "_UCellMedianMaxRank")]]
```
## plotting
### Main 5f boxplot by BH4 score (primary only)
```{r}
# 1. Build the long-format dataframe for plotting (signatures only)
box_long <- do.call(rbind, lapply(bh4_score_names, function(s) {
  data.frame(
    Quantity = s,
    cancer_state = seu$cell_state,
    Origin = seu$Origin,
    value = as.numeric(seu@meta.data[[paste0(s, "_UCellMedianMaxRank")]]),
    y_label = "Signature score",
    stringsAsFactors = FALSE
  )
})) %>% subset(cancer_state != "Other" & Origin %in% c("Pm0", "Pm1"))

# 2. Generate and save the boxplot using your custom function
title_str <- sprintf("BH4 signatures by cell state (primary only) | UCell (maxRank = %d)", ucell_maxrank_custom)
out_file  <- file.path(current.outputfolder, "5f_Boxplot_byCellState_BH4_score.pdf")

make_boxplot_state(
  box_long,
  bh4_score_names,
  title_str,
  out_file,
  levels = STATE_ORDER[1:3],
  fills = state_color[1:3]
)
```

## write session info
```{r session-info}
sessionInfo()
writeLines(capture.output(sessionInfo()), "PDAC_human_park_BH4_scoring_sessionInfo.txt")
```
