---
title: "TCGA BRCA RNAseq - EMT and Gene Set Scoring (BRCA subset, rescored)"
author: "Charlie Whittaker"
output:
  html_document:
    toc: true
    toc_depth: 3
  html_notebook: default
date: "8/11/2026"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(future.globals.maxSize = 5e10)
set.seed(1234)
```

# Purpose

`TCGA_EMT_GeneSet_Scoring.Rmd` scores EMT (KS) and gene sets (GSVA,
ssGSEA, UCell) once, pan-cancer, across all ~11,069 TCGA samples. This
notebook picks up that already-scored object, subsets it down to BRCA
only (~1,215 samples), and re-runs all scoring methods on the BRCA subset
alone. Every sample here is BRCA, so the KS score isn't given a
distinguishing suffix -- it's recomputed under its original column name
and simply overwrites the inherited pan-cancer values; GSVA/ssGSEA/UCell
are (re)scored against `pathways_to_score` only (see below), so those
columns are new regardless.

**Why this isn't a no-op for every method** -- the scoring methods differ
in whether they reference the rest of the cohort at all:

- **KSScore()** -- a per-sample KS test comparing that one sample's own
  Epi-gene values against its own Mes-gene values. Never looks at other
  samples, so BRCA-only and pan-cancer-context KS scores for the same
  sample are expected to be numerically identical.
- **GSVA/ssGSEA** -- even with `kcdf = "none"`, each gene's expression is
  converted to a rank/ECDF value computed *across all samples in the
  input matrix*. A smaller, BRCA-only background is a different
  reference population, so these scores are expected to shift.
- **UCell** -- ranks genes *within each sample* only, never across
  samples, so like KS it's expected to be numerically identical whether
  run pan-cancer or BRCA-only.

```{r load-packages, message=FALSE, warning=FALSE}
library(tidyverse)
library(openxlsx)
library(ggpubr)
library(Seurat)
library(GSVA)
library(UCell)
library(ComplexHeatmap)
```

# Load TCGA BRCA object

```{r load-and-subset}
tcga_brca  <- readRDS("TCGA_BRCA_seurat.rds")
```

# Log-transform

`CreateSeuratObject()` only populates the `counts` layer, not `data` --
this creates `data` directly from `counts` (log2-transformed) rather than
going through `NormalizeData()`, whose per-sample total-count division is
the extra normalization step decided against above.

```{r log-transform}
DefaultAssay(tcga_brca) <- "RNA"
tcga_brca <- SetAssayData(
  tcga_brca, layer = "data",
  new.data = log2(GetAssayData(tcga_brca, layer = "counts") + 1)
)
# if `layer` isn't accepted for this assay class/Seurat version, use:
# slot = "data" / slot = "counts" instead
```

# Point to ks signature file and read in gene sets

```{r ks-signature-genes}
KSScore.EMTSignature.file <- "../emt_code/EM_gene_signature_tumor_KS.xlsx"

emt.tumor <- read.xlsx("../emt_code/EM_gene_signature_tumor_KS.xlsx", colNames = FALSE)
emt.tumor <- emt.tumor %>% dplyr::rename(gene = X1, group = X2)
Epi.tum <- emt.tumor$gene[emt.tumor$group == "Epi"]
Mes.tum <- emt.tumor$gene[emt.tumor$group == "Mes"]
```

# Pathway gene sets for GSVA scoring

```{r}
# BH4 synthesis and recycling
BH4 <- c("DHFR", "GCH1", "PCBD1", "PCBD2", "PTS", "QDPR", "SPR")

# De novo BH4 synthesis only
BH4.synth <- c("GCH1", "PTS", "SPR")

# BH4 recycling and regeneration only
BH4.recyc <- c("QDPR", "DHFR", "PCBD1", "PCBD2")

# Cystine uptake and GSH synthesis/recycling
Glutathione <- c("SLC7A11", "SLC3A2", "GCLC", "GCLM", "GSS", "GSR")

# CoQ/Ubiquinol biosynthesis
CoQ10 <- c("COQ2", "COQ3", "COQ4", "COQ5", "COQ6", "COQ7",
            "COQ8A", "COQ8B", "COQ9", "HPDL", "PDSS1", "PDSS2", "STARD7")

# Thioredoxin system
Thioredoxin <- c("TXN", "TXNRD1", "TXN2", "TXNRD2")

# Canonical peroxiredoxins
Peroxiredoxin <- c("PRDX1", "PRDX2", "PRDX3", "PRDX4", "PRDX5")

# Individual genes
goi <- c("GPX4", "AIFM2", "DHODH", "PRDX6")
```

## Create a list of these gene sets, except for GOI

```{r}
pathways_to_score <- list(
  BH4           = BH4,
  BH4.synth     = BH4.synth,
  BH4.recyc     = BH4.recyc,
  Glutathione   = Glutathione,
  CoQ10         = CoQ10,
  Thioredoxin   = Thioredoxin,
  Peroxiredoxin = Peroxiredoxin,
  Epi.tum       = Epi.tum,
  Mes.tum       = Mes.tum)
```

# Re-score EMT on the BRCA subset

## KS method (Chakraborty/Bell)

```{r source-ks-method}
source("../emt_code/Chakraborty_et_al_KS_code_vCAW.R")
```

```{r rescore-ks}
emt_genes <- unique(c(Epi.tum, Mes.tum))

expr_matrix_brca <- GetAssayData(tcga_brca, layer = "data")
expr_matrix_brca <- expr_matrix_brca[rownames(expr_matrix_brca) %in% emt_genes, ]
expr_matrix_brca <- as.matrix(expr_matrix_brca)
dim(expr_matrix_brca)

ks_tumor_brca <- suppressWarnings(KSScore(expr_matrix_brca, "TCGA_BRCAonly_tumorSig", "."))

tcga_brca$EMT_KS_TumorSig <- ks_tumor_brca[colnames(tcga_brca)]
```

# Score pathway gene sets (BH4/redox panel + Epi.tum/Mes.tum)

## GSVA

```{r score-gsva}
expr_matrix_full_brca <- as.matrix(GetAssayData(tcga_brca, layer = "data"))
dim(expr_matrix_full_brca)

gsva_param_brca <- gsvaParam(expr_matrix_full_brca, pathways_to_score, kcdf = "none")
gsva_scores_brca <- gsva(gsva_param_brca)
dim(gsva_scores_brca)  # expect 9 gene sets x ncol(tcga_brca)

gsva_scores_brca_df <- as.data.frame(t(gsva_scores_brca))
colnames(gsva_scores_brca_df) <- paste0(colnames(gsva_scores_brca_df), "_GSVA")

tcga_brca <- AddMetaData(tcga_brca, gsva_scores_brca_df)
```

## ssGSEA

Reuses `expr_matrix_full_brca` and `pathways_to_score` from the GSVA step
above, scored with `ssgseaParam()` instead of `gsvaParam()`.

```{r score-ssgsea}
ssgsea_param_brca <- ssgseaParam(expr_matrix_full_brca, pathways_to_score, normalize = TRUE)
ssgsea_scores_brca <- gsva(ssgsea_param_brca)
dim(ssgsea_scores_brca)  # expect 9 gene sets x ncol(tcga_brca)

ssgsea_scores_brca_df <- as.data.frame(t(ssgsea_scores_brca))
colnames(ssgsea_scores_brca_df) <- paste0(colnames(ssgsea_scores_brca_df), "_ssGSEA")

tcga_brca <- AddMetaData(tcga_brca, ssgsea_scores_brca_df)
```

## UCell

Reuses `pathways_to_score` from above, scored directly on `tcga_brca`'s
`data` layer with `AddModuleScore_UCell()` (`maxRank = 10000`).

```{r score-ucell}
tcga_brca <- AddModuleScore_UCell(
  tcga_brca, features = pathways_to_score, maxRank = 10000,
  slot = "data", name = "_UCell"
)

ucell_cols_brca <- grep("_UCell$", colnames(tcga_brca@meta.data), value = TRUE)
ucell_cols_brca
```

# Save re-scored BRCA object

```{r save-brca-scored}
saveRDS(tcga_brca, "TCGA_BRCA_seurat_EMTscored.rds")
```

# KS tertiles, based on the new BRCA scores

Same tertile logic as `Atlas_05`/`TCGA_EMT_GeneSet_Scoring.Rmd`, applied
to the newly-computed BRCA KS scores, so the visualizations below reflect
BRCA-calibrated groupings.

```{r ks-tertile-group}
.ksTertileGroup <- function(x) {
  ks_tertiles <- quantile(x, probs = c(1/3, 2/3), na.rm = TRUE)
  cut(x, breaks = c(-Inf, ks_tertiles[1], ks_tertiles[2], Inf),
      labels = c("KS_low", "KS_mid", "KS_high"))
}

tcga_brca$KS_Group <- .ksTertileGroup(tcga_brca$EMT_KS_TumorSig)

table(tcga_brca$KS_Group)
```

# Compare GSVA vs ssGSEA scoring

Same 9 gene sets (`pathways_to_score`), GSVA score vs ssGSEA score for
each, one panel per gene set.

```{r gsva-vs-ssgsea-data}
gsva_ssgsea_long <- purrr::map_dfr(names(pathways_to_score), function(gs) {
  tibble(
    GeneSet = gs,
    GSVA    = tcga_brca@meta.data[[paste0(gs, "_GSVA")]],
    ssGSEA  = tcga_brca@meta.data[[paste0(gs, "_ssGSEA")]]
  )
}) %>%
  mutate(GeneSet = factor(GeneSet, levels = names(pathways_to_score)))
```

```{r gsva-vs-ssgsea-scatter, fig.width=10, fig.height=10}
ggplot(gsva_ssgsea_long, aes(x = GSVA, y = ssGSEA)) +
  geom_point(size = 0.6, alpha = 0.4) +
  geom_smooth(method = "lm", color = "red", linewidth = 0.6, se = FALSE) +
  stat_cor(size = 3) +
  facet_wrap(~ GeneSet, ncol = 3, scales = "free") +
  theme_bw() +
  labs(x = "GSVA score", y = "ssGSEA score",
       title = "GSVA vs ssGSEA score, by gene set (TCGA BRCA)")
```

# BH4 (+ subsets) score by KS tertile

Same decomposition as `Atlas_05`/`TCGA_EMT_GeneSet_Scoring.Rmd`'s BH4 box
plots, using the BRCA GSVA scores and KS tertile groups.

```{r bh4-long-scores-fn}
.bh4LongScores <- function(object, group_var) {
  score_vars <- c("BH4_GSVA", "BH4.synth_GSVA", "BH4.recyc_GSVA")

  FetchData(object, vars = c(score_vars, group_var)) %>%
    pivot_longer(cols = all_of(score_vars), names_to = "col_name", values_to = "Score") %>%
    dplyr::mutate(
      GeneSet = factor(sub("_GSVA$", "", col_name), levels = c("BH4", "BH4.synth", "BH4.recyc"))
    )
}
```

```{r bh4-boxplot-by-ksgroup, fig.width=13, fig.height=4}
bh4_ks_long <- .bh4LongScores(tcga_brca, "KS_Group")
bh4_ks_group_n <- table(tcga_brca$KS_Group)

ggplot(bh4_ks_long, aes(x = KS_Group, y = Score, fill = KS_Group)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.8, linewidth = 0.7) +
  geom_jitter(width = 0.2, alpha = 0.3, size = 0.8) +
  facet_wrap(~ GeneSet, ncol = 3, scales = "free_y") +
  theme_bw() +
  theme(
    panel.grid   = element_blank(),
    panel.border = element_rect(colour = "black", fill = NA, linewidth = 1.2),
    axis.title   = element_text(face = "bold", size = 13),
    axis.text    = element_text(face = "bold", size = 11)
  ) +
  NoLegend() +
  scale_x_discrete(labels = function(l) paste0(l, "\n(n=", bh4_ks_group_n[l], ")")) +
  labs(x = NULL, y = "GSVA score")
```

## BH4 (GSVA) score vs EMT KS score

```{r bh4-vs-ks-scatter, fig.width=5, fig.height=5}
ggplot(tcga_brca@meta.data, aes(x = EMT_KS_TumorSig, y = BH4_GSVA)) +
  geom_point(size = 0.8, alpha = 0.5) +
  geom_smooth(method = "lm", color = "red", linewidth = 0.6, se = FALSE) +
  stat_cor(size = 3) +
  theme_bw() +
  labs(x = "EMT KS score", y = "BH4 score (GSVA)",
       title = "BH4 score (GSVA) vs EMT KS score")
```

## BH4 gene expression vs EMT KS score

Raw expression (not the gene-set score) for each of BH4's 7 genes,
individually, against the BRCA KS score -- `facet_wrap(~ Gene, ncol = 2)`
lays 7 genes out as 4 rows x 2 columns (last cell empty).

```{r bh4-genes-vs-emt-data}
bh4_gene_expr_long <- FetchData(tcga_brca, vars = c(BH4, "EMT_KS_TumorSig")) %>%
  pivot_longer(cols = all_of(BH4), names_to = "Gene", values_to = "Expression")
```

```{r bh4-genes-vs-ks-scatter, fig.width=8, fig.height=12}
ggplot(bh4_gene_expr_long, aes(x = EMT_KS_TumorSig, y = Expression)) +
  geom_point(size = 0.6, alpha = 0.4) +
  geom_smooth(method = "lm", color = "red", linewidth = 0.6, se = FALSE) +
  stat_cor(size = 3) +
  facet_wrap(~ Gene, ncol = 2, scales = "free_y") +
  theme_bw() +
  labs(x = "EMT KS score", y = "Log-normalized expression",
       title = "BH4 gene expression vs EMT KS score")
```

# Manual dot plots

Same manual mean-expression/pct-expressed dot plot as
`TCGA_EMT_GeneSet_Scoring.Rmd`/`Atlas_05`. `type` isn't a useful grouping
here (every sample is BRCA), so this uses `KS_Group` instead.

```{r manual-dotplot-fn}
.manualDotPlotData <- function(object, genes, group_var, assay = "RNA", layer = "data") {
  expr_data <- GetAssayData(object, assay = assay, layer = layer)

  missing_genes <- setdiff(genes, rownames(expr_data))
  if (length(missing_genes) > 0) {
    stop("Gene(s) not found in ", assay, "/", layer, ": ", paste(missing_genes, collapse = ", "))
  }
  expr_data <- expr_data[genes, , drop = FALSE]

  groups <- object@meta.data[[group_var]]
  group_levels <- if (is.factor(groups)) levels(groups) else sort(unique(groups))

  dot_long <- purrr::map_dfr(group_levels, function(g) {
    # which(), not `groups == g` directly -- groups may contain NA, and NA
    # in a logical index produces an NA element in the result rather than
    # being dropped, which then errors out sparse-Matrix column indexing
    cells_in_group <- colnames(object)[which(groups == g)]
    expr_g <- expr_data[, cells_in_group, drop = FALSE]
    tibble(
      gene    = genes,
      group   = g,
      meanExp = Matrix::rowMeans(expr_g),
      pctExp  = Matrix::rowMeans(expr_g > 0) * 100
    )
  })

  dot_long %>%
    dplyr::group_by(gene) %>%
    dplyr::mutate(
      zscore  = as.numeric(scale(meanExp)),
      # min-max normalized per gene, 0-1 -- z-score gets unstable/exaggerated
      # with very few groups (e.g. 3 KS tertiles: scale() on n=3 points can
      # look extreme), so this is the steadier option for small group
      # counts. Not meaningful with 1 group (denominator is 0 -> NaN).
      normExp = (meanExp - min(meanExp)) / (max(meanExp) - min(meanExp))
    ) %>%
    dplyr::ungroup() %>%
    dplyr::mutate(
      gene  = factor(gene, levels = genes),
      group = factor(group, levels = group_levels)
    )
}

.manualDotPlot <- function(dot_long, x_lab = "Gene", y_lab = "Group",
                            color_by = c("zscore", "normExp")) {
  color_by <- match.arg(color_by)

  if (color_by == "zscore") {
    lim <- max(abs(dot_long$zscore), na.rm = TRUE)
    lim <- ifelse(lim == 0, 1, lim)

    p <- ggplot(dot_long, aes(x = gene, y = group)) +
      geom_point(aes(size = pctExp, color = zscore)) +
      scale_color_gradientn(
        colors = c("#2F6FB2", "#9FC3DF", "#F4EFE6", "#F28E5C", "#E84A3A"),
        values = scales::rescale(c(-lim, -lim / 3, 0, lim / 3, lim)),
        limits = c(-lim, lim),
        name   = "Z-score\nmean expr."
      )
  } else {
    p <- ggplot(dot_long, aes(x = gene, y = group)) +
      geom_point(aes(size = pctExp, color = normExp)) +
      scale_color_gradientn(
        colors = c("#2F6FB2", "#9FC3DF", "#F4EFE6", "#F28E5C", "#E84A3A"),
        limits = c(0, 1),
        name   = "Normalized\nmean expr."
      )
  }

  p +
    scale_size_continuous(name = "% expressed", range = c(0.5, 10)) +
    theme_minimal(base_size = 10) +
    theme(
      axis.text.x  = element_text(angle = 45, hjust = 1, face = "italic"),
      panel.grid   = element_blank(),
      panel.border = element_rect(colour = "grey40", fill = NA, linewidth = 0.6)
    ) +
    labs(x = x_lab, y = y_lab)
}
```

## BH4 genes: dot plot by KS tertile

```{r dotplot-bh4-by-ksgroup, fig.width=8, fig.height=5}
bh4_dot_ksgroup <- .manualDotPlotData(tcga_brca, BH4, "KS_Group")
.manualDotPlot(bh4_dot_ksgroup, x_lab = "Gene", y_lab = "KS tertile", color_by = "normExp")
```

# Gene-set score heatmap

Same heatmap strategy as `TCGA_EMT_GeneSet_Scoring.Rmd`/`Atlas_05` --
9 already-scored gene-set columns plus `goi`'s 4 genes' raw expression,
samples ordered by `EMT_KS_TumorSig`, rows in a fixed order
(`Epi.tum`/`Mes.tum` right below the color bar, then the pathway gene
sets in their existing order, then `goi`), z-scored per row, with
`EMT_KS_TumorSig` as a continuous color bar, and a row-side strip
distinguishing gene-set-score rows from `goi`'s raw-expression rows.

```{r genesetscore-heatmap-fn}
.geneSetScoreHeatmap <- function(object, suffix, title) {
  score_set_order <- c("Epi.tum", "Mes.tum", "BH4", "BH4.synth", "BH4.recyc",
                        "Glutathione", "CoQ10", "Thioredoxin", "Peroxiredoxin")
  score_cols_hm <- paste0(score_set_order, suffix)

  ks_order  <- order(object@meta.data$EMT_KS_TumorSig)
  ks_values <- object@meta.data$EMT_KS_TumorSig[ks_order]

  score_mat <- t(as.matrix(FetchData(object, vars = score_cols_hm)))
  score_mat <- score_mat[, ks_order]
  rownames(score_mat) <- sub(suffix, "", rownames(score_mat))

  goi_expr <- GetAssayData(object, assay = "RNA", layer = "data")[goi, , drop = FALSE]
  goi_mat  <- as.matrix(goi_expr)[, ks_order, drop = FALSE]

  combined_mat    <- rbind(score_mat, goi_mat)
  combined_scaled <- t(scale(t(combined_mat)))

  ks_cols <- circlize::colorRamp2(c(min(ks_values), max(ks_values)), c("darkorange", "lightgreen"))

  col_anno <- HeatmapAnnotation(
    EMT_KS_TumorSig = ks_values,
    col = list(EMT_KS_TumorSig = ks_cols)
  )

  row_type <- factor(
    c(rep("Gene set score", length(score_set_order)), rep("Gene (goi)", length(goi))),
    levels = c("Gene set score", "Gene (goi)")
  )
  row_anno <- rowAnnotation(
    `Row type` = row_type,
    col = list(`Row type` = c("Gene set score" = "#666666", "Gene (goi)" = "#F2A104")),
    show_annotation_name = FALSE
  )

  Heatmap(combined_scaled,
          name              = "z-score",
          top_annotation    = col_anno,
          left_annotation   = row_anno,
          cluster_columns   = FALSE,
          cluster_rows      = FALSE,
          show_column_names = FALSE,
          column_title      = paste0(title, ", ordered by EMT_KS_TumorSig (Epi -> Mes)"),
          row_title         = "Gene set score / gene expression")
}

score_cols_hm_suffix_gsva <- "_GSVA"
```

```{r heatmap-gsva, fig.width=11, fig.height=8}
.geneSetScoreHeatmap(tcga_brca, score_cols_hm_suffix_gsva, "TCGA BRCA (GSVA)")
```

# Requested Figures

Same single-panel boxplot design as `TCGA_EMT_GeneSet_Scoring.Rmd`/
`Atlas_05` -- one gene set at a time, one panel, tight whisker-based
y-axis, and every pairwise comparison shown (not just the non-significant
ones), with black brackets and BH-adjusted p-value labels. No omnibus
Kruskal-Wallis annotation: with ~1,215 BRCA samples split 3 ways, it comes
back significant on essentially every panel regardless of effect size, and
it isn't gating anything here (the pairwise brackets render either way) --
so it added a second, less specific p-value rather than new information.
Parameterized by `method` (`"GSVA"`/`"ssGSEA"`/`"UCell"`) so the same
functions drive all three sets of figures below. Different, more targeted
view than the existing "BH4 (+ subsets) score by KS tertile" section
above (which covers BH4/BH4.synth/BH4.recyc together, no significance
testing) -- not a replacement for it.

```{r bh4-single-panel-fn}
.tightBoxYlim <- function(df, y_col, group_col) {
  group_ranges <- tapply(df[[y_col]], df[[group_col]], function(y) {
    range(boxplot.stats(y)$stats)
  })
  c(min(sapply(group_ranges, min)), max(sapply(group_ranges, max)))
}

# All pairwise Wilcoxon tests among group_col's levels, BH-adjusted --
# returns EVERY pair, tagged `significant` (p.adj < 0.05), with the
# BH-adjusted p-value itself as the label (not significance stars). Pairs
# with too few samples in either group are silently skipped rather than
# erroring.
.bh4PairwiseTests <- function(df, y_col, group_col, min_n = 2) {
  groups <- levels(df[[group_col]])
  pairs  <- utils::combn(groups, 2, simplify = FALSE)

  res <- purrr::map_dfr(pairs, function(pr) {
    x <- df[[y_col]][df[[group_col]] == pr[1]]
    y <- df[[y_col]][df[[group_col]] == pr[2]]
    if (length(x) < min_n || length(y) < min_n) return(tibble())
    tibble(group1 = pr[1], group2 = pr[2], p = wilcox.test(x, y)$p.value)
  })

  if (nrow(res) == 0) return(res)

  res$p.adj       <- p.adjust(res$p, method = "BH")
  res$significant <- res$p.adj < 0.05
  res$p.label     <- paste0("p = ", formatC(res$p.adj, format = "g", digits = 2))
  res
}

.bh4SinglePanel <- function(object, group_var, x_lab, method = c("GSVA", "ssGSEA", "UCell"),
                             add_pairwise_brackets = TRUE) {
  method    <- match.arg(method)
  score_col <- paste0("BH4_", method)

  df <- FetchData(object, vars = c(score_col, group_var))
  colnames(df)[colnames(df) == score_col] <- "Score"
  colnames(df)[colnames(df) == group_var]  <- "Group"

  ylim_vals <- .tightBoxYlim(df, "Score", "Group")

  pairwise <- NULL
  if (add_pairwise_brackets) {
    pairwise <- .bh4PairwiseTests(df, "Score", "Group")
    if (nrow(pairwise) > 0) {
      # brackets stack above the box tops -- extend the tight range just
      # enough to fit however many are actually being drawn
      step <- 0.12 * diff(ylim_vals)
      pairwise$y.position <- ylim_vals[2] + seq_len(nrow(pairwise)) * step
      ylim_vals[2] <- max(pairwise$y.position) + step * 0.5
    }
  }

  group_n <- table(df$Group)

  p <- ggplot(df, aes(x = Group, y = Score, fill = Group)) +
    geom_boxplot(outlier.shape = NA, alpha = 0.8, linewidth = 0.7) +
    coord_cartesian(ylim = ylim_vals) +
    theme_bw() +
    theme(
      panel.grid   = element_blank(),
      panel.border = element_rect(colour = "black", fill = NA, linewidth = 1.2),
      axis.title   = element_text(face = "bold", size = 13),
      axis.text    = element_text(face = "bold", size = 11)
    ) +
    NoLegend() +
    scale_x_discrete(labels = function(l) paste0(l, "\n(n=", group_n[l], ")")) +
    labs(x = NULL, y = paste0(method, " score"))

  if (add_pairwise_brackets && !is.null(pairwise) && nrow(pairwise) > 0) {
    p <- p + ggpubr::stat_pvalue_manual(
      pairwise, label = "p.label", tip.length = 0.01,
      color = "black", fontface = "bold", bracket.size = 0.6
    )
  }

  p
}
```

```{r bh4-gsva-boxplot-by-ksgroup, fig.width=4, fig.height=3}
.bh4SinglePanel(tcga_brca, "KS_Group", "KS tertile", method = "GSVA")
```

```{r bh4-ssgsea-boxplot-by-ksgroup, fig.width=5, fig.height=5}
fig.5c <- .bh4SinglePanel(tcga_brca, "KS_Group", "KS tertile", method = "ssGSEA")

ggsave("figures/Fig5c_ssgsea_BH4_TCGA_BRCA.svg", plot = fig.5c, width = 5, height = 5, dpi = 300)

fig.5c
```

```{r bh4-ucell-boxplot-by-ksgroup, fig.width=4, fig.height=3}
.bh4SinglePanel(tcga_brca, "KS_Group", "KS tertile", method = "UCell")
```

## Same single-panel design, any gene set

Generalizes `.bh4SinglePanel()` to take a `gene_set` argument instead of
being hardcoded to BH4 -- everything else (tight whisker-based y-axis,
all-pairwise brackets in black, the GSVA/ssGSEA/UCell `method` switch) is
identical. Reuses `.tightBoxYlim()`/`.bh4PairwiseTests()` from
`bh4-single-panel-fn` above as-is, since neither one actually references
BH4 specifically.

```{r genesetscore-single-panel-fn}
.geneSetSinglePanel <- function(object, gene_set, group_var, x_lab,
                                 method = c("GSVA", "ssGSEA", "UCell"),
                                 add_pairwise_brackets = TRUE) {
  method    <- match.arg(method)
  score_col <- paste0(gene_set, "_", method)

  df <- FetchData(object, vars = c(score_col, group_var))
  colnames(df)[colnames(df) == score_col] <- "Score"
  colnames(df)[colnames(df) == group_var] <- "Group"

  ylim_vals <- .tightBoxYlim(df, "Score", "Group")

  pairwise <- NULL
  if (add_pairwise_brackets) {
    pairwise <- .bh4PairwiseTests(df, "Score", "Group")
    if (nrow(pairwise) > 0) {
      step <- 0.12 * diff(ylim_vals)
      pairwise$y.position <- ylim_vals[2] + seq_len(nrow(pairwise)) * step
      ylim_vals[2] <- max(pairwise$y.position) + step * 0.5
    }
  }

  group_n <- table(df$Group)

  p <- ggplot(df, aes(x = Group, y = Score, fill = Group)) +
    geom_boxplot(outlier.shape = NA, alpha = 0.8, linewidth = 0.7) +
    coord_cartesian(ylim = ylim_vals) +
    theme_bw() +
    theme(
      panel.grid   = element_blank(),
      panel.border = element_rect(colour = "black", fill = NA, linewidth = 1.2),
      axis.title   = element_text(face = "bold", size = 13),
      axis.text    = element_text(face = "bold", size = 11)
    ) +
    NoLegend() +
    scale_x_discrete(labels = function(l) paste0(l, "\n(n=", group_n[l], ")")) +
    labs(x = NULL, y = paste0(method, " score"))

  if (add_pairwise_brackets && !is.null(pairwise) && nrow(pairwise) > 0) {
    p <- p + ggpubr::stat_pvalue_manual(
      pairwise, label = "p.label", tip.length = 0.01,
      color = "black", fontface = "bold", bracket.size = 0.6
    )
  }

  p
}
```

Gene sets under test, beyond BH4 -- add/remove names here to change what
the loops below produce; each just needs `<name>_GSVA`/`<name>_ssGSEA`/
`<name>_UCell` columns to already exist on `tcga_brca` (see
`pathways_to_score`).

```{r requested-gene-sets}
requested_gene_sets <- c("Glutathione", "CoQ10", "Thioredoxin", "Peroxiredoxin")
```

```{r genesets-gsva-boxplot-by-ksgroup, fig.width=4, fig.height=3}
for (gs in requested_gene_sets) {
  print(.geneSetSinglePanel(tcga_brca, gs, "KS_Group", "KS tertile", method = "GSVA"))
}
```

```{r genesets-ssgsea-boxplot-by-ksgroup, fig.width=4, fig.height=3}
for (gs in requested_gene_sets) {
  print(.geneSetSinglePanel(tcga_brca, gs, "KS_Group", "KS tertile", method = "ssGSEA"))
}
```

```{r genesets-ucell-boxplot-by-ksgroup, fig.width=4, fig.height=3}
for (gs in requested_gene_sets) {
  print(.geneSetSinglePanel(tcga_brca, gs, "KS_Group", "KS tertile", method = "UCell"))
}
```

## Gene-set score heatmap (no goi rows)

Same simplification just applied to Atlas_05's `genesetscore-heatmap-fn`
-- 7 gene-set-score rows only (`Epi.tum`/`Mes.tum`/`BH4`/`Glutathione`/
`CoQ10`/`Thioredoxin`/`Peroxiredoxin`; `BH4.synth`/`BH4.recyc` and `goi`'s
raw expression dropped), samples ordered by `EMT_KS_TumorSig`, z-scored
per row, with `EMT_KS_TumorSig` as the only color bar. A more targeted
view than the existing "Gene-set score heatmap" section above (9 rows +
goi) -- not a replacement for it. One heatmap per scoring method, same
`suffix`-driven function as before.

```{r genesetscore-heatmap-requested-fn}
.geneSetScoreHeatmapRequested <- function(object, suffix, title) {
  score_set_order <- c("Epi.tum", "Mes.tum", "BH4",
                        "Glutathione", "CoQ10", "Thioredoxin", "Peroxiredoxin")
  score_cols_hm <- paste0(score_set_order, suffix)

  ks_order  <- order(object@meta.data$EMT_KS_TumorSig)
  ks_values <- object@meta.data$EMT_KS_TumorSig[ks_order]

  score_mat <- t(as.matrix(FetchData(object, vars = score_cols_hm)))
  score_mat <- score_mat[, ks_order]
  rownames(score_mat) <- sub(suffix, "", rownames(score_mat))

  combined_scaled <- t(scale(t(score_mat)))

  ks_cols <- circlize::colorRamp2(c(min(ks_values), max(ks_values)), c("darkorange", "lightgreen"))

  col_anno <- HeatmapAnnotation(
    EMT_KS_TumorSig = ks_values,
    col = list(EMT_KS_TumorSig = ks_cols)
  )

  Heatmap(combined_scaled,
          name              = "z-score",
          top_annotation    = col_anno,
          cluster_columns   = FALSE,
          cluster_rows      = FALSE,
          show_column_names = FALSE,
          column_title      = paste0(title, ", ordered by EMT_KS_TumorSig (Epi -> Mes)"),
          row_title         = "Gene set score")
}

score_cols_hm_suffix_ssgsea <- "_ssGSEA"
score_cols_hm_suffix_ucell  <- "_UCell"
```

```{r heatmap-gsva-requested, fig.width=11, fig.height=6}
.geneSetScoreHeatmapRequested(tcga_brca, score_cols_hm_suffix_gsva, "TCGA BRCA (GSVA)")
```

```{r heatmap-ssgsea-requested, fig.width=11, fig.height=6}
.geneSetScoreHeatmapRequested(tcga_brca, score_cols_hm_suffix_ssgsea, "TCGA BRCA (ssGSEA)")
```

```{r heatmap-ucell-requested, fig.width=11, fig.height=6}
.geneSetScoreHeatmapRequested(tcga_brca, score_cols_hm_suffix_ucell, "TCGA BRCA (UCell)")
```

# write session info

Capturing information about the R session helps document your work

```{r session-info}
sessionInfo()
writeLines(capture.output(sessionInfo()), "TCGA_BRCA_EMT_GeneSet_Scoring_sessionInfo.txt")
```
