---
title: "Whitney Henry - scRNAseq Analysis - CZ CELLxGENE Atlas - Malignant Pseudobulk Prep"
author: "Charlie Whittaker"
output:
  html_document:
    toc: true
    toc_depth: 3
  html_notebook: default
date: "9/10/2026"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(future.globals.maxSize = 5e10)
set.seed(1234)
```

# Goal
This workflow does not need to be run but instead documents the preparation of the necessary pseudobulk object.

`ChenAtlas_malignant_pseudobulk.Rmd` starts from `Atlas_malignant_pseudobulk_donor.rds`
and only actually needs one thing from it: the per-donor expression matrix
(plus donor metadata) -- it computes `EMT_KS_TumorSig` (used for `KS_Group`
tertiles) and recomputes GSVA/ssGSEA/UCell itself, all directly in that
notebook.

This notebook is a single, lightweight run that regenerates exactly that
object.

The source h5ad is downloaded on demand into `data/` (see below) rather
than assumed to already be there, so this notebook is runnable from a clean
checkout with nothing but the CZ CELLxGENE URL.

Nothing is checkpointed to disk along the way -- the ~621K-cell full atlas
and the malignant-only subset (multi-GB Seurat objects in the original
pipeline) exist only in this session's memory and are dropped (`rm()`/`gc()`)
once no longer needed. Only the final donor-level pseudobulk object (tens of
MB) is written out. The tradeoff: there's no restart checkpoint, so a crash
partway through means re-running from the top, including the full h5ad
import and gene-symbol resolution over ~621K cells.

```{r load-packages, message=FALSE, warning=FALSE}
library(tidyverse)
library(Matrix)
library(Seurat)
library(anndataR)
```

# Import h5ad and resolve to gene-symbol-keyed Seurat object

Same source atlas, layer-naming fix, and Ensembl->symbol resolution. The max-per-gene (MPG) approach is 
used to resolve genes that share a symbol.

```{r data-paths}
h5ad_url  <- "https://datasets.cellxgene.cziscience.com/0fbf6fa3-8463-46ab-bda8-9703848b58db.h5ad"
h5ad_file <- "data/0fbf6fa3-8463-46ab-bda8-9703848b58db.h5ad"
```

Downloads the source atlas into `data/` if it isn't already there (this
is a ~1-2GB file)

```{r download-h5ad}
dir.create(dirname(h5ad_file), recursive = TRUE, showWarnings = FALSE)
if (!file.exists(h5ad_file)) {
  download.file(h5ad_url, destfile = h5ad_file, mode = "wb")
}
```

```{r import-h5ad}
seu <- read_h5ad(h5ad_file, as = "Seurat")
seu
```

```{r rename-X-layer}
# `X` is log-normalized data (confirmed in Atlas_01_import.Rmd), not raw
# counts -- belongs in Seurat's `data` slot, not `counts`.
seu[["RNA"]]$data <- seu[["RNA"]]$X
seu[["RNA"]]$X <- NULL
```

```{r source-mpg-tool}
source("Rcode/data_formatting_tools.R")
```

```{r compute-gene-mpg}
adata <- read_h5ad(h5ad_file)  # default `as`, for the var/feature_name table

mat <- GetAssayData(seu, assay = "RNA", layer = "data")
matched_var <- adata$var[match(rownames(mat), rownames(adata$var)), ]

mpg_table <- data.frame(
  ensembl_id   = rownames(mat),
  feature_name = as.character(matched_var$feature_name),
  Avg          = Matrix::rowMeans(mat),
  Var          = Matrix::rowMeans(mat^2) - Matrix::rowMeans(mat)^2,
  stringsAsFactors = FALSE
)

mpg_table <- calculate_MPG(mpg_table, gene_col = "feature_name", avg_col = "Avg", var_col = "Var")
rm(adata, mat)
```

```{r rename-features-to-symbol}
mat_mpg <- GetAssayData(seu, assay = "RNA", layer = "data")
mat_mpg <- mat_mpg[mpg_table$MPG == "Yes", ]
rownames(mat_mpg) <- mpg_table$feature_name[mpg_table$MPG == "Yes"]

seu[["RNA"]] <- CreateAssay5Object(data = mat_mpg)
rm(mat_mpg, mpg_table)

dim(seu)  # one row per unique gene symbol now, not one per Ensembl ID
```

No further cell filtering is applied here (no mito/QC metrics ship with this
export, and none of the 8 batches stood out on `nCount_RNA`/`nFeature_RNA` --
see `Atlas_01_import.Rmd`'s QC section for that check) -- proceeds on the
full imported object as-is.

# Subset to malignant cells

```{r subset-malignant}
DefaultAssay(seu) <- "RNA"
malignant <- subset(seu, subset = cell_type == "malignant cell")
malignant

rm(seu)
gc()
```

# Aggregate to donor-level pseudobulk (malignant cells only)

`AverageExpression()`, not `AggregateExpression()`, since this atlas has no
raw counts anywhere to sum -- it un-logs each log-normalized value, means
within donor, then re-logs (see `Atlas_02_allcells_pseudobulk.Rmd`/`Atlas_04`
for the full reasoning). `data` from this call is already properly
re-logged, so no further `NormalizeData()` step is needed.

```{r aggregate-pseudobulk}
pseudobulk <- AverageExpression(
  malignant,
  assays        = "RNA",
  layer         = "data",
  group.by      = "donor_id",
  return.seurat = TRUE
)
# if `layer` isn't accepted for this Seurat/AverageExpression version, use:
# slot = "data" instead

pseudobulk
```

## Reattach donor-level clinical metadata

`AverageExpression()` sanitizes `donor_id` into column names (`_` -> `-`),
so the join below matches on the sanitized form and verifies that recovers
every column before trusting it -- same fix as `Atlas_02`/`Atlas_04`. `assay`
is excluded: it's a per-library technical detail (donor `BC397_Tumor` has
cells under two different chemistries), not a stable per-donor field.

```{r reattach-metadata}
donor_meta <- malignant@meta.data %>%
  dplyr::distinct(donor_id, batch, disease, tissue, sex,
                   self_reported_ethnicity, development_stage, grade)

stopifnot(!any(duplicated(donor_meta$donor_id)))

rownames(donor_meta) <- gsub("_", "-", donor_meta$donor_id)
stopifnot(all(colnames(pseudobulk) %in% rownames(donor_meta)))

donor_meta_ordered <- donor_meta[colnames(pseudobulk), ]

pseudobulk$donor_id <- donor_meta_ordered$donor_id
pseudobulk@meta.data <- cbind(
  pseudobulk@meta.data,
  donor_meta_ordered[, setdiff(colnames(donor_meta_ordered), "donor_id")]
)
```

## Add malignant cell counts per donor

```{r add-donor-counts}
donor_counts <- table(malignant$donor_id)
donor_counts_df <- as.data.frame(donor_counts, stringsAsFactors = FALSE)
colnames(donor_counts_df) <- c("donor_id", "n_malignant_cells")

rownames(donor_counts_df) <- gsub("_", "-", donor_counts_df$donor_id)
stopifnot(all(colnames(pseudobulk) %in% rownames(donor_counts_df)))

pseudobulk$n_malignant_cells <- donor_counts_df[colnames(pseudobulk), "n_malignant_cells"]

pseudobulk@meta.data

rm(malignant)
gc()
```

# Save pseudobulk object

No EMT/KS scoring is done here -- `ChenAtlas_malignant_pseudobulk.Rmd` now
computes `EMT_KS_TumorSig` itself directly from this object's expression
matrix, so this checkpoint carries only the aggregated expression matrix
and donor metadata/counts from the sections above.

```{r save-pseudobulk}
saveRDS(pseudobulk, "Atlas_malignant_pseudobulk_donor.rds")
```

# write session info

```{r session-info}
sessionInfo()
writeLines(capture.output(sessionInfo()), "ChenAtlas_malignant_pseudobulk_prep_sessionInfo.txt")
```
