---
title: "Whitney Henry - scRNAseq Analysis - CZ CELLxGENE Atlas - Malignant Pseudobulk ssGSEA/GSVA/UCell Figures"
author: "Charlie Whittaker"
output:
  html_document:
    toc: true
    toc_depth: 3
  html_notebook: default
date: "8/31/2026"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
set.seed(1234)
```

```{r load-packages, message=FALSE, warning=FALSE}
library(tidyverse)
library(openxlsx)
library(ggpubr)
library(Seurat)
library(GSVA)
library(UCell)
library(ComplexHeatmap)
```

# Goal

This is the decided-on figure set for the malignant pseudobulk data:
gene-set scores by `EMT_KS_TumorSig` tertile, as box plots with statistics,
for all three scoring methods on `pseudobulk` -- ssGSEA, GSVA, and UCell --
plus a single ssGSEA summary heatmap. EMT itself is scored with the KS
method only (MLR is not used in this notebook); samples are stratified
into low/mid/high tertiles of that KS score.

It starts from `Atlas_malignant_pseudobulk_donor.rds`, produced by
`ChenAtlas_malignant_pseudobulk_prep.Rmd`. That checkpoint carries only
the per-donor expression matrix and donor metadata -- no score columns --
so this notebook is fully self-contained.

# Load pseudobulk object

```{r load-pseudobulk}
pseudobulk <- readRDS("Atlas_malignant_pseudobulk_donor.rds")
pseudobulk
```

# EMT KS scoring

`Atlas_malignant_pseudobulk_donor.rds` carries only the per-donor
expression matrix and donor metadata -- `EMT_KS_TumorSig` itself is
computed here using the `KSScore()` method to run a per-donor, two-sample
Kolmogorov-Smirnov test comparing each donor's own expression across the
tumor-calibrated `Mes.tum` genes against its own expression across the
`Epi.tum` genes.

```{r ks-source}
source("emt_code/Chakraborty_et_al_KS_code_vCAW.R")
KSScore.EMTSignature.file <- "emt_code/EM_gene_signature_tumor_KS.xlsx"

emt.tumor.ks <- read.xlsx("emt_code/EM_gene_signature_tumor_KS.xlsx", colNames = FALSE)
emt.tumor.ks <- emt.tumor.ks %>% dplyr::rename(gene = X1, group = X2)
Epi.tum.ks <- emt.tumor.ks$gene[emt.tumor.ks$group == "Epi"]
Mes.tum.ks <- emt.tumor.ks$gene[emt.tumor.ks$group == "Mes"]
```

```{r ks-tumor-pseudobulk}
emt_genes_ks <- unique(c(Epi.tum.ks, Mes.tum.ks))

expr_matrix_ks <- GetAssayData(pseudobulk, assay = "RNA", layer = "data")
expr_matrix_ks <- expr_matrix_ks[rownames(expr_matrix_ks) %in% emt_genes_ks, ]
expr_matrix_ks <- as.matrix(expr_matrix_ks)
dim(expr_matrix_ks)

ks_tumor_pb <- suppressWarnings(KSScore(expr_matrix_ks, "tumorSig_AtlasMalignantPseudobulk", "."))
pseudobulk$EMT_KS_TumorSig <- ks_tumor_pb[colnames(pseudobulk)]
```

# KS score tertiles

Same tertile logic used across the companion bulk/pseudobulk notebooks
(bottom/middle/top third of `EMT_KS_TumorSig`), applied to the
`EMT_KS_TumorSig` score just computed above.

```{r ks-tertile-fn}
.ksTertileGroup <- function(x) {
  ks_tertiles <- quantile(x, probs = c(1/3, 2/3), na.rm = TRUE)
  cut(x, breaks = c(-Inf, ks_tertiles[1], ks_tertiles[2], Inf),
      labels = c("KS_low", "KS_mid", "KS_high"))
}

pseudobulk$KS_Group <- .ksTertileGroup(pseudobulk$EMT_KS_TumorSig)

table(pseudobulk$KS_Group)
```

# Score gene sets with GSVA, ssGSEA, and UCell

`Atlas_04` already scored `pseudobulk` with GSVA and ssGSEA; all three
methods are (re-)scored here instead, on the exact same 9-gene-set list --
GSVA/ssGSEA reuse `Atlas_04`'s own `gsvaParam()`/`ssgseaParam()` calls
verbatim (this overwrites the `_GSVA`/`_ssGSEA` columns already on
`pseudobulk` -- expected, since the input matrix and gene sets haven't
changed) and UCell is scored fresh, since `Atlas_04` doesn't compute it.
The pathway/KS-signature gene sets are rebuilt here -- identical
definitions to `Atlas_04`'s own `pathways_to_score`.

```{r ucell-gene-sets}
# BH4 synthesis and recycling
BH4 <- c("DHFR", "GCH1", "PCBD1", "PCBD2", "PTS", "QDPR", "SPR")

# De novo BH4 synthesis only
BH4.synth <- c("GCH1", "PTS", "SPR")

# BH4 recycling and regeneration only
BH4.recyc <- c("QDPR", "DHFR", "PCBD1", "PCBD2")

# Cystine uptake and GSH synthesis/recycling
Glutathione <- c("SLC7A11", "SLC3A2", "GCLC", "GCLM", "GSS", "GSR")

# CoQ/Ubiquinol biosynthesis
CoQ10 <- c("COQ2", "COQ3", "COQ4", "COQ5", "COQ6", "COQ7",
            "COQ8A", "COQ8B", "COQ9", "HPDL", "PDSS1", "PDSS2", "STARD7")

# Thioredoxin system
Thioredoxin <- c("TXN", "TXNRD1", "TXN2", "TXNRD2")

# Canonical peroxiredoxins
Peroxiredoxin <- c("PRDX1", "PRDX2", "PRDX3", "PRDX4", "PRDX5")

emt.tumor <- read.xlsx("emt_code/EM_gene_signature_tumor_KS.xlsx", colNames = FALSE)
emt.tumor <- emt.tumor %>% dplyr::rename(gene = X1, group = X2)
Epi.tum <- emt.tumor$gene[emt.tumor$group == "Epi"]
Mes.tum <- emt.tumor$gene[emt.tumor$group == "Mes"]

pathways_to_score <- list(
  BH4           = BH4,
  BH4.synth     = BH4.synth,
  BH4.recyc     = BH4.recyc,
  Glutathione   = Glutathione,
  CoQ10         = CoQ10,
  Thioredoxin   = Thioredoxin,
  Peroxiredoxin = Peroxiredoxin,
  Epi.tum       = Epi.tum,
  Mes.tum       = Mes.tum)
```

## GSVA

Same call as `Atlas_04`'s `gsva-pseudobulk`/`store-gsva-scores` chunks --
no `kcdf = "none"` override here (unlike the bulk RNA-seq notebooks), to
match `Atlas_04` exactly.

```{r gsva-pseudobulk}
expr_matrix_pb <- as.matrix(GetAssayData(pseudobulk, assay = "RNA", layer = "data"))
dim(expr_matrix_pb)

gsva_param  <- gsvaParam(expr_matrix_pb, pathways_to_score)
gsva_scores <- gsva(gsva_param)
dim(gsva_scores)  # expect 9 gene sets x ncol(pseudobulk)

gsva_scores_df <- as.data.frame(t(gsva_scores))
colnames(gsva_scores_df) <- paste0(colnames(gsva_scores_df), "_GSVA")

pseudobulk <- AddMetaData(pseudobulk, gsva_scores_df)
```

## ssGSEA

Same call as `Atlas_04`'s `ssgsea-pseudobulk`/`store-ssgsea-scores`
chunks. Reuses `expr_matrix_pb` and `pathways_to_score` from the GSVA step
above.

```{r ssgsea-pseudobulk}
ssgsea_param  <- ssgseaParam(expr_matrix_pb, pathways_to_score, normalize = TRUE)
ssgsea_scores <- gsva(ssgsea_param)
dim(ssgsea_scores)  # expect 9 gene sets x ncol(pseudobulk)

ssgsea_scores_df <- as.data.frame(t(ssgsea_scores))
colnames(ssgsea_scores_df) <- paste0(colnames(ssgsea_scores_df), "_ssGSEA")

pseudobulk <- AddMetaData(pseudobulk, ssgsea_scores_df)
```

## UCell

```{r ucell-pseudobulk}
pseudobulk <- UCell::AddModuleScore_UCell(
  pseudobulk, features = pathways_to_score, maxRank = 10000,
  slot = "data", name = "_UCell"
)

ucell_cols_pb <- grep("_UCell$", colnames(pseudobulk@meta.data), value = TRUE)
ucell_cols_pb
```

# Gene sets scored

Pulled directly from the `_ssGSEA` columns the scoring section above
attached to `pseudobulk`'s metadata, rather than re-deriving the list from
`pathways_to_score` -- confirms the actual scored columns, not just the
gene sets that were requested.

`_GSVA` and `_UCell` are expected to cover the exact same set of gene sets
(all three were scored from the same `pathways_to_score` above) -- checked
explicitly below rather than assumed, so a mismatched gene-set list fails
loudly here instead of silently producing empty/partial plots further
down.

```{r scored-gene-sets}
ssgsea_cols <- grep("_ssGSEA$", colnames(pseudobulk@meta.data), value = TRUE)
gsva_cols   <- grep("_GSVA$",   colnames(pseudobulk@meta.data), value = TRUE)
ucell_cols  <- grep("_UCell$",  colnames(pseudobulk@meta.data), value = TRUE)

all_scored_gene_sets <- sub("_ssGSEA$", "", ssgsea_cols)

stopifnot(
  "GSVA-scored gene sets don't match ssGSEA's -- check the GSVA scoring chunk above" =
    setequal(all_scored_gene_sets, sub("_GSVA$", "", gsva_cols)),
  "UCell-scored gene sets don't match ssGSEA's -- check the UCell scoring chunk above" =
    setequal(all_scored_gene_sets, sub("_UCell$", "", ucell_cols))
)

all_scored_gene_sets
```

# Box plots by KS tertile, one section per scoring method

## Shared helpers

`.tightBoxYlim()` computes each KS tertile's own boxplot whisker range
(`boxplot.stats()`, the same lower/upper-whisker definition `geom_boxplot()`
uses) and takes the min/max across the three tertiles, then
`coord_cartesian()` zooms to exactly that -- tight to the boxes themselves,
not stretched by outlier donors that aren't even drawn.
`.pairwiseTests()` runs all pairwise Wilcoxon tests among `KS_Group`'s
levels, BH-adjusted for the 3 comparisons, and returns every pair tagged
`significant` (p.adj < 0.05) with the BH-adjusted p-value itself shown as
the label (not significance stars) -- both significant and
non-significant pairs are shown on the plot.

```{r boxplot-helper-fns}
.tightBoxYlim <- function(df, y_col, group_col) {
  group_ranges <- tapply(df[[y_col]], df[[group_col]], function(y) {
    range(boxplot.stats(y)$stats)
  })
  c(min(sapply(group_ranges, min)), max(sapply(group_ranges, max)))
}

.pairwiseTests <- function(df, y_col, group_col, min_n = 2) {
  groups <- levels(df[[group_col]])
  pairs  <- utils::combn(groups, 2, simplify = FALSE)

  res <- purrr::map_dfr(pairs, function(pr) {
    x <- df[[y_col]][df[[group_col]] == pr[1]]
    y <- df[[y_col]][df[[group_col]] == pr[2]]
    if (length(x) < min_n || length(y) < min_n) return(tibble())
    tibble(group1 = pr[1], group2 = pr[2], p = wilcox.test(x, y)$p.value)
  })

  if (nrow(res) == 0) return(res)

  res$p.adj       <- p.adjust(res$p, method = "BH")
  res$significant <- res$p.adj < 0.05
  res$p.label     <- paste0("p = ", formatC(res$p.adj, format = "g", digits = 2))
  res
}
```

## Single-panel plotting function

One panel per gene set: box plot by `KS_Group` and all-pairwise Wilcoxon
brackets (black, BH-adjusted p-value labels). No omnibus Kruskal-Wallis annotation --
dropped for consistency with `TCGA_BRCA_EMT_GeneSet_Scoring.Rmd`/
`TCGA_EMT_GeneSet_Scoring.Rmd`: it isn't gating anything here (the brackets
render regardless of its result), and the BH-adjusted pairwise comparisons
are the more specific, more informative answer to "which tertiles differ"
that these panels are actually being used for.

```{r genesetscore-single-panel-fn}
.geneSetSinglePanelByKS <- function(object, gene_set, suffix, method_label,
                                     add_pairwise_brackets = TRUE) {
  score_col <- paste0(gene_set, suffix)

  df <- FetchData(object, vars = c(score_col, "KS_Group"))
  colnames(df)[colnames(df) == score_col] <- "Score"

  ylim_vals <- .tightBoxYlim(df, "Score", "KS_Group")

  pairwise <- NULL
  if (add_pairwise_brackets) {
    pairwise <- .pairwiseTests(df, "Score", "KS_Group")
    if (nrow(pairwise) > 0) {
      step <- 0.12 * diff(ylim_vals)
      pairwise$y.position <- ylim_vals[2] + seq_len(nrow(pairwise)) * step
      ylim_vals[2] <- max(pairwise$y.position) + step * 0.5
    }
  }

  group_n <- table(df$KS_Group)

  p <- ggplot(df, aes(x = KS_Group, y = Score, fill = KS_Group)) +
    geom_boxplot(outlier.shape = NA, alpha = 0.8, linewidth = 0.7) +
    coord_cartesian(ylim = ylim_vals) +
    theme_bw() +
    theme(
      panel.grid   = element_blank(),
      panel.border = element_rect(colour = "black", fill = NA, linewidth = 1.2),
      plot.title   = element_text(face = "bold", size = 13, hjust = 0.5),
      axis.title   = element_text(face = "bold", size = 13),
      axis.text    = element_text(face = "bold", size = 11)
    ) +
    NoLegend() +
    scale_x_discrete(labels = function(l) paste0(l, "\n(n=", group_n[l], ")")) +
    labs(x = NULL, y = paste0(method_label, " score"),
         title = paste0(gene_set, " (", method_label, ")"))

  if (add_pairwise_brackets && !is.null(pairwise) && nrow(pairwise) > 0) {
    p <- p + ggpubr::stat_pvalue_manual(
      pairwise, label = "p.label", tip.length = 0.01,
      color = "black", fontface = "bold", bracket.size = 0.6
    )
  }

  p
}
```

## ssGSEA: one box plot per gene set

Only the 5 gene sets below are saved to `figures/` -- these are the
specifically-named SVGs pulled into the paper; every other gene set is
printed inline only, nothing else is written to disk.

```{r geneset-boxplot-ksgroup-ssgsea, fig.width=4, fig.height=3}
required_svg_figures <- c(
  BH4           = "figures/Fig5b_ssgsea_BH4_Chen.svg",
  Glutathione   = "figures/ExtData5a_ssgsea_Glutathione_Chen.svg",
  CoQ10         = "figures/ExtData5b_ssgsea_CoQ10_Chen.svg",
  Thioredoxin   = "figures/ExtData5c_ssgsea_Thioredoxin_Chen.svg",
  Peroxiredoxin = "figures/ExtData5d_ssgsea_Peroxiredoxin_Chen.svg"
)

for (gs in all_scored_gene_sets) {
  p <- .geneSetSinglePanelByKS(pseudobulk, gs, "_ssGSEA", "ssGSEA")

  if (gs %in% names(required_svg_figures)) {
    ggsave(required_svg_figures[[gs]], plot = p, width = 4, height = 3)
    ggsave(sub("\\.svg$", ".pdf", required_svg_figures[[gs]]), plot = p, width = 4, height = 3)
  }

  print(p)
}
```

## GSVA: one box plot per gene set

Same `.geneSetSinglePanelByKS()` call as ssGSEA above, just pointed at the
`_GSVA` columns scored above -- same tight-whisker y-axis and BH-adjusted
pairwise Wilcoxon brackets, one panel per gene set.
Printed inline only -- none of the requested figures are GSVA-based, so
nothing here is saved to `figures/`.

```{r geneset-boxplot-ksgroup-gsva, fig.width=4, fig.height=3}
for (gs in all_scored_gene_sets) {
  p <- .geneSetSinglePanelByKS(pseudobulk, gs, "_GSVA", "GSVA")

  print(p)
}
```

## UCell: one box plot per gene set

Same again, pointed at the `_UCell` columns scored above (directly on
`pseudobulk`, `maxRank = 10000`). Printed inline only, nothing saved to
`figures/`.

```{r geneset-boxplot-ksgroup-ucell, fig.width=4, fig.height=3}
for (gs in all_scored_gene_sets) {
  p <- .geneSetSinglePanelByKS(pseudobulk, gs, "_UCell", "UCell")

  print(p)
}
```

# Heatmap: ssGSEA scores across donors, ordered by KS score

Donors ordered by `EMT_KS_TumorSig` (Epi -> Mes), with the KS score itself
shown as a top color bar. Rows are the 7 curated gene-set scores (`BH4.synth`/
`BH4.recyc`, subsets of `BH4`'s own genes, are excluded here), each z-scored
across donors since the gene sets aren't on a shared scale to begin with.

```{r genesetscore-heatmap-fn}
.geneSetScoreHeatmap <- function(object, suffix, title) {
  score_set_order <- c("Epi.tum", "Mes.tum", "BH4",
                        "Glutathione", "CoQ10", "Thioredoxin", "Peroxiredoxin")
  score_cols_hm <- paste0(score_set_order, suffix)

  ks_order  <- order(object@meta.data$EMT_KS_TumorSig)
  ks_values <- object@meta.data$EMT_KS_TumorSig[ks_order]

  score_mat <- t(as.matrix(FetchData(object, vars = score_cols_hm)))
  score_mat <- score_mat[, ks_order]
  rownames(score_mat) <- sub(suffix, "", rownames(score_mat))

  combined_scaled <- t(scale(t(score_mat)))

  ks_cols <- circlize::colorRamp2(c(min(ks_values), max(ks_values)), c("darkorange", "lightgreen"))

  col_anno <- HeatmapAnnotation(
    EMT_KS_TumorSig = ks_values,
    col = list(EMT_KS_TumorSig = ks_cols)
    )

  Heatmap(combined_scaled,
          name              = "z-score",
          top_annotation    = col_anno,
          cluster_columns   = FALSE,
          cluster_rows      = FALSE,
          show_column_names = FALSE,
          column_title      = paste0(title, ", ordered by EMT_KS_TumorSig (Epi -> Mes)"),
          row_title         = "Gene set score")
}
```

```{r pb-heatmap-ssgsea, fig.width=13, fig.height=6}
.geneSetScoreHeatmap(pseudobulk, "_ssGSEA", "Pseudobulk (ssGSEA)")
```

# write session info

Capturing information about the R session helps document your work

```{r session-info}
sessionInfo()
writeLines(capture.output(sessionInfo()), "ChenAtlas_pseudobulk_figures_sessionInfo.txt")
```
