---
title: "Whitney Henry - Bulk RNAseq Analysis - Neuroblastoma Treehouse (PMID 40603900, GSE294351) - BH4 vs MES-ADRN Correlation"
author: "Evelyn Tong"
output:
  html_document:
    toc: true
    toc_depth: 3
  html_notebook: default
date: "9/22/2026"
---

Split out of the original monolithic `tmp/WH_BH4pathway_final/src/Analysis_script.Rmd`
("BH4 analysis script - Evelyn - 2026/09/16", Data6 section) during the
`datasets/` reorg, into its own self-contained dataset directory/project.
See `NB_Treehouse_BH4_MESADRN_correlation_Methods.md` for the full
methodology. This notebook is nearly identical to the companion SEQC
notebook (`Neuroblastoma_human_25150838_SEQC_GSE62564/`) -- same
methodology, different cohort.

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, message = FALSE, warning = FALSE,
  fig.align = "center", fig.width = 6, fig.height = 4, dpi = 300
)
source("emt_code/bh4_pathway_shared_setup.R")
```

## parameter settings
```{r}
# figures/ + stats output for this dataset
current.outputfolder <- "figures"
# define where the input RDS file is
data.path <- "Treehouse_NB.rds"
```
## load data and processing
```{r}
seu <- readRDS(data.path)
gene_seu  <- seu$gene_matrix
clinical_seu <- seu$clinical_metadata

cat("  Dimensions:", dim(gene_seu)[1], "genes x", dim(gene_seu)[2], "cells\n")
cat("  BH4 genes present:", sum(bh4_genes %in% rownames(gene_seu)), "/", length(bh4_genes), "\n")
```
## calculate ADRN/MES and BH4 signature scores (ssgsea)
```{r}
mesadrn.signatures <- readxl::read_excel("28650485_Groningen_SuppTable2.xlsx", skip = 2, col_names = FALSE)
colnames(mesadrn.signatures) <- c("gene", "state")
pathways_to_score <- c(list(
  MES     = mesadrn.signatures[mesadrn.signatures$state == "MES", "gene"]$gene,
  ADRN = mesadrn.signatures[mesadrn.signatures$state == "ADRN", "gene"]$gene),
  bh4_sigs)

# Calculate normalized ssGSEA scores directly on the gene_seu matrix
# (modern GSVA param-object API -- see BRCA_TCGA/SKCM_TCGA notebooks;
# GSVA >= 2.x removed the old gsva(matrix, sets, method = ...) call)
ssgsea_param <- ssgseaParam(as.matrix(gene_seu), pathways_to_score, normalize = TRUE)
ssgsea_mat <- gsva(ssgsea_param)
# Convert the resulting (pathway x sample) matrix to a dataframe
scores_ssgsea <- as.data.frame(t(ssgsea_mat))

# Add sample IDs and calculate the derived relative scores
scores_ssgsea$sample_id <- rownames(scores_ssgsea)
scores_ssgsea$ADRN_minus_MES <- scores_ssgsea$ADRN - scores_ssgsea$MES
scores_ssgsea$MES_ADRN_score <- scores_ssgsea$ADRN_minus_MES * -1   # = MES - ADRN

# Optional: Reorder columns for a clean output format
scores_ssgsea <- scores_ssgsea[, c("sample_id", names(pathways_to_score), "ADRN_minus_MES", "MES_ADRN_score")]

scores_ssgsea$BH4_score <- scores_ssgsea[[bh4_full]]
```
## run spearman correlation - BH4 Signatures vs MES-ADRN score
```{r}
# Calculate correlation for each signature in bh4_score_names
corr_rows <- lapply(bh4_score_names, function(sig) {
  ct <- suppressWarnings(
    cor.test(scores_ssgsea$MES_ADRN_score, scores_ssgsea[[sig]], method = "spearman")
  )

  data.frame(
    Signature = sig,
    Spearman_rho = unname(ct$estimate),
    p_value = ct$p.value,
    stringsAsFactors = FALSE
  )
})

# Combine into a single dataframe
corr_df <- do.call(rbind, corr_rows)

# Apply Benjamini-Hochberg (FDR) correction for multiple testing
corr_df$p_value_BH <- p.adjust(corr_df$p_value, method = "BH")
corr_df$significant <- corr_df$p_value_BH < 0.05

# Save the statistics to a CSV file
out_file <- "BH4_signatures_spearman_correlation_stats_ssgsea.csv"
write.csv(corr_df, out_file, row.names = FALSE)
```
## plotting
### Main 5i scatter plot
```{r}
on_color <- "#FF9400"
# Create a list of the 3 scatter plots
panels <- lapply(bh4_score_names, function(sig) {

  # Extract correlation stats for this signature from the previously created corr_df
  ct_row <- corr_df[corr_df$Signature == sig, ]
  p_str  <- if (ct_row$p_value < 0.001) sprintf("p = %.2e", ct_row$p_value) else sprintf("p = %.3f", ct_row$p_value)
  subtitle_str <- sprintf("rho = %.3f, %s", ct_row$Spearman_rho, p_str)

  # Setup plotting dataframe
  df_plot <- data.frame(
    x = scores_ssgsea$MES_ADRN_score,
    y = scores_ssgsea[[sig]]
  )

  # Calculate linear fit
  fit <- lm(y ~ x, data = df_plot)
  df_plot$fit <- predict(fit)

  # Generate single plot
  ggplot(df_plot, aes(x = x, y = y)) +
    geom_point(size = 0.4, alpha = 0.4, color = "black") +
    geom_line(aes(y = fit), color = on_color, linewidth = 0.5) +
    labs(
      title = sig,
      subtitle = subtitle_str,
      x = "MES-ADRN score",
      y = "BH4 score"
    ) +
    figure_theme +
    coord_fixed(ratio = diff(range(df_plot$x, na.rm = TRUE)) / diff(range(df_plot$y, na.rm = TRUE)))
})

# Combine the 3 panels into 1 row, 3 columns using patchwork
p_scatter <- wrap_plots(panels, ncol = 3) +
  plot_annotation(
    title = sprintf("BH4 Signatures vs MES-ADRN Score (Treehouse dataset, n=%d) | ssGSEA", nrow(scores_ssgsea)),
    theme = theme(plot.title = element_text(face = "bold", size = 6, hjust = 0.5))
  )

# Save the figure (both PDF and SVG)
out_file <- file.path(current.outputfolder, "5i_scatterplot_BH4vsMESADRN_score.pdf")
pdf(out_file, width = 5, height = 2)  # Wider layout for 3 columns in 1 row
print(p_scatter)
dev.off()
svg(sub("\\.pdf$", ".svg", out_file), width = 5, height = 2)
print(p_scatter)
dev.off()

# Display in the RStudio session / notebook output (the pdf()/svg() device
# blocks above only draw to their respective files, not the screen)
print(p_scatter)
```

## write session info
```{r session-info}
sessionInfo()
writeLines(capture.output(sessionInfo()), "NB_Treehouse_BH4_MESADRN_correlation_sessionInfo.txt")
```
